---
title: "Adding plot boundaries to BETYdb"
author: "Kimberly Huynh"
date: "5/20/2019"
output: html_document
---

# Objective

The objective of this tutorial is to demonstrate how to add plot boundaries to the BETYdb sites table. For this tutorial, data will be added to a local instance of a test database. Instructions on how to set up a test database can be found [here](https://gist.github.com/kimberlyh66/935106e35478829b9e844ba5b496784d).

Shapefiles for MAC Season 8 sites will be used as an example.

# Set up

Load in the `sf` and `RPostgreSQL` packages.

```{r load-pack, message = FALSE}

library(sf)
library(RPostgreSQL)

```

Create a connection to the BETY test database.

```{r create-dbcon}

dbcon <- dbConnect(PostgreSQL(),
                   host = 'localhost',
                   user = 'bety',
                   port = 5433)

```

# Adding plot boundaries using `sf` package

## Read in shapefiles using `st_read`

MAC Season 8 shapefiles can be read in as a `simple features object` using `st_read`. These shapefiles are saved in the `MAC_season8` folder of this directory. 

A `sf` object is a dataframe where rows are features (points, lines, polygons) and columns are attributes of the features.

```{r read-shape, results = 'hide'}

s8_bounds <- st_read('MAC_season8/S8_one_row_polys.shp')

```

If you need to convert the coordinate reference system, you can use the `st_transform` function. In this example, we will be converting the EPSG from `32612` to `4326`.

```{r transform-crs}

s8_bounds <- st_transform(s8_bounds, 4326)

```

## Write a table to BETYdb using `st_write`

The `st_write` function can be used to write a new table to BETYdb containing the contents of the `sf` object. The new table will be named `s8_bounds` and will be dropped after it is used to insert new data into the `sites` table.

```{r write-table}

st_write(s8_bounds, dbcon)

```

# Updating and inserting new records using `RPostgreSQL` package

The MAC Season 8 shapefiles did not inlude a `sitename` attribute. We will need to add this field to the `s8_bounds` table and update its values.

The `dbSendStatement` function from the `RPostgreSQL` package will be used to execute SQL statements. 

The following statements will be executed:

`alter table s8_bounds add column sitename VARCHAR;`

`update s8_bounds set sitename = concat('MAC Field Scanner Season 8 Range ', f_range, ' Column ', f_column, ' ', row);`

```{r create-update-sitename, results = 'hide'}

# create sitename field in s8_bounds table
create_sitename <- dbSendStatement(dbcon,
                                   "alter table s8_bounds add column sitename VARCHAR;")
dbClearResult(create_sitename)

# populate sitename values to be concatenation of multiple fields from table
update_sitename <- dbSendStatement(dbcon,
                                   paste0("update s8_bounds ",
                                          "set sitename = concat(",
                                          "'MAC Field Scanner Season 8 Range ', ",
                                          "f_range, ",
                                          "' Column ', ",
                                          "f_column, ",
                                          "' ', ",
                                          "row);"))
dbClearResult(update_sitename)

```

## Add new records to sites table using subset of s8_bounds

The `sitename` and `geometry` fields of the `s8_bounds` table will be selected to insert new records into the sites table. You may need to use functions such as `st_force3D` or `st_transform` to transform the `geometry` field. 

SQL statement to insert new records into sites: 

`insert into sites (sitename, geometry) select sitename, st_force3D(geometry) from s8_bounds;`

```{r insert-sites, results = 'hide'}

# insert new records into sites table using records from s8_bounds
insert_sites <- dbSendStatement(dbcon,
                                paste0("insert into sites (sitename, geometry) ",
                                       "select sitename, st_force3D(geometry) from s8_bounds;"))
dbClearResult(insert_sites)

```

## Drop table

Once the new MAC season 8 plot records have been added to the sites table, the `s8_bounds` table should be dropped.

SQL statement to drop table: 

`drop table s8_bounds;`

```{r drop-s8-bounds, results = 'hide'}

# drop the s8_bounds table
drop_table <- dbSendStatement(dbcon, "drop table s8_bounds;")
dbClearResult(drop_table)

```

# Close database connection

```{r close-dbcon, results = 'hide'}

dbDisconnect(dbcon)

```

