---
title: 'Metadata Upload Tutorial Part 2: Associating metadata in BETYdb'
author: "Kimberly Huynh"
date: "4/30/2019"
output: html_document
---

# Overview

The objective of this tutorial is to demonstrate how to associate the metadata that you have added to the traits database. Each site that you add must be associated with an experiment, cultivar, and citation. Each treatment that you add must be associated with an experiment. 

This tutorial should be completed after the `Metadata Upload Tutorial Part 1: Adding new metadata to BETYdb`.

For this tutorial, make sure that you have a local instance of a test database running and that all new experiments, sites, treatments, cultivars, and citations have been added.

# How to run this Rmd

Knitting this markdown file will not associate any data in the test database. Being able to knit this file means that you will not encounter any errors when associating experiments with sites, experiments with treatments, sites with cultivars, and citations with sites.

To commit changes to the test database, you will need to make an edit to this markdown file. 

`commit = TRUE` needs to be set in the following chunk.

```{r commit-param}

commit <- FALSE

```

# Set up

## URL (required input)

Please provide the URL to your google sheet below. 

_this should be the same URL that you used for the first tutorial_

A sample URL has been provided for a google sheet containing MAC season 8 metadata. Replace the sample URL with your own.

```{r url}

url <- 'https://docs.google.com/spreadsheets/d/1BPkZPfynZI9DB9bRJfa335uC_uNjST5jHVf-XUgSVvM/edit#gid=347205052' 

```

## Load packages 

The `metadata` package will need to be loaded along with `RPostgres`, `dplyr`, `kableExtra`, `googlesheets` and `glue`. 

```{r load-pack, message = FALSE}

library(metadata)
library(RPostgres)
library(readxl)
library(dplyr)
library(kableExtra)
library(googlesheets)
library(glue)

```

## Read in google sheets

Read in the users, sites, treatments, and citations sheets. 

```{r read-user-gs, message = FALSE}

# get key from URL
key <- extract_key_from_url(url)

# register sheet 
gs_obj <- key %>% gs_key(lookup = FALSE)

# users sheet
users <- gs_obj %>% gs_read(ws = 'users')

# sites sheet
sites <- gs_obj %>% gs_read(ws = 'sites')

# treatments sheet
treatments <- gs_obj %>% gs_read(ws = 'treatments')

# citations sheet
citations <- gs_obj %>% gs_read(ws = 'citations')

```

# Metadata association

## Connect to database

A connection to the test database needs to be created. Make sure that you have the test database running and your new metadata has been added.

The port number should be the same as the one indicated in docker-compose.override.yml.

```{r create-dbcon}

dbcon <- dbConnect(RPostgres::Postgres(),
                   host = 'localhost',
                   user = 'bety',
                   port = 5433)

```

## Functions to be used for metadata association

You will be using the following functions from the `metadata` package to associate your data: 

   * `associate_experiments_sites`
   
   * `associate_experiments_treatments`
   
   * `associate_sites_cultivars`
   
   * `associate_citations_sites`

## Step 1: Associate experiments with sites

To associate experiments with sites, run `associate_experiments_sites`. The sites google sheet object will need to be passed in. 

```{r associate-exp-site, results = 'hide'}

associate_experiments_sites(dbcon, commit, sites)

```

## Step 2: Associate experiments with treatments

To associate experiments with treatments, run `associate_experiments_treatments`. The treatments google sheet object will need to be passed in.

```{r associate-exp-treat, results = 'hide'}

associate_experiments_treatments(dbcon, commit, treatments)

```

## Step 3: Associate sites with cultivars

To associate sites with cultivars, run `associate_sites_cultivars`. The sites google sheet object will need to be passed in.

```{r associate-site-cultivar, results = 'hide'}

associate_sites_cultivars(dbcon, commit, sites)

```

## Step 4: Associate citations with sites

To associate citations with sites, run `associate_citations_sites`. The citations and sites google sheet objects will need to be passed in.

```{r associate-citation-site, results = 'hide'}

associate_citations_sites(dbcon, commit, citations, sites)

```

# Check data association

You can check if your data has been correctly associated in the database once you run this markdown file with `commit <- TRUE`.

These chunks will only be evaluated if `commit <- TRUE`. If evaluated, you should see a table of fetched records below a chunk.

We will be using your user id and today's date to fetch the data you uploaded.

```{r today}

today <- Sys.Date()

# get user id from user login
bety_users <- tbl(dbcon, 'users') %>% collect()
  
user_id <- bety_users %>%
  filter(login == users$login) %>%
  select(id)

```

## Fetch records from experiments_sites

```{r fetch-exp-site, eval = commit}

exp_site_query <- dbSendQuery(dbcon,
                              glue_sql("select * from experiments_sites where experiment_id in 
                                       (select id from experiments where created_at::date = {today} and 
                                       user_id = (select id from users where login = {users$login})) and 
                                       site_id in (select id from sites where created_at::date = {today} and
                                       user_id = (select id from users where login = {users$login})) limit 10",
                                       .con = dbcon))

exp_site_fetch <- dbFetch(exp_site_query)
dbClearResult(exp_site_query)

kable(exp_site_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

## Fetch records from experiments_treatments

```{r fetch-exp-treat, eval = commit}

exp_treat_query <- dbSendQuery(dbcon,
                               glue_sql("select * from experiments_treatments where experiment_id in
                                        (select id from experiments where created_at::date = {today} and
                                        user_id = (select id from users where login = {users$login})) and                                                  treatment_id in (select id from treatments where created_at::date = {today}                                         and user_id = (select id from users where login = {users$login}))",
                                        .con = dbcon))

exp_treat_fetch <- dbFetch(exp_treat_query)
dbClearResult(exp_treat_query)

kable(exp_treat_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

## Fetch records from sites_cultivars

```{r fetch-site-cultivar, eval = commit}

site_cultivar_query <- dbSendQuery(dbcon,
                                   glue_sql("select * from sites_cultivars where site_id in
                                            (select id from sites where created_at::date = {today} and 
                                            user_id = (select id from users where login = {users$login})) 
                                            limit 10",
                                            .con = dbcon))

site_cultivar_fetch <- dbFetch(site_cultivar_query)
dbClearResult(site_cultivar_query)

kable(site_cultivar_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

## Fetch records from citations_sites

```{r fetch-citation-site, eval = commit}

citation_site_query <- dbSendQuery(dbcon,
                                   glue_sql("select * from citations_sites where citation_id in
                                            (select id from citations where created_at::date = {today} and 
                                            user_id = (select id from users where login = {users$login})) and 
                                            site_id in (select id from sites where created_at::date = {today} and
                                            user_id = (select id from users where login = {users$login})) 
                                            limit 10",
                                            .con = dbcon))

citation_site_fetch <- dbFetch(citation_site_query)
dbClearResult(citation_site_query)

kable(citation_site_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

## Close database connection

```{r close-dbcon}

dbDisconnect(dbcon)

```
