---
title: 'Metadata Upload Tutorial Part 2: Associations'
author: "Kimberly Huynh"
date: "4/30/2019"
output: html_document
---

# Overview

The objective of this tutorial is to demonstrate how to associate metadata that you have added to the traits database. 

This tutorial should be completed after `Metadata Upload Tutorial Part 1: Additions`.

For this tutorial, make sure that you have a local instance of a test database running and that all new experiments, sites, treatments, cultivars, and citations have been added.

## Notes on Running this Rmd

Knitting this markdown file will not associate any data in the test database. Being able to knit this file means that you will not encounter any errors when associating experiments with sites, experiments with treatments, sites with cultivars, and citations with sites.

To make associations in the test database, you will need to make an edit to this markdown file. 

`commit = TRUE` needs to be set in the following chunk.

```{r commit-param}

commit <- FALSE

```

## Check associated data

You will not be able to fetch any of the data you associate until you have commit your changes to the database. 

You can check your upload results when you run this Rmd with `commit <- TRUE` in the above chunk. 

The value of `eval_chk` will depend on the value you set for `commit`. This variable will determine whether chunks for fetching uploaded data will be evaluated. You will not see any tables below these chunks unless you set commit to TRUE.  

```{r check-upload, results = 'hide'}

ifelse(commit == TRUE, eval_chk <- TRUE, eval_chk <- FALSE)

```

# Required input

URL to a public published google sheet (please provide in set up section)

_this should be the same URL that you used for the `Metadata Upload Tutorial Part 1: Addition` tutorial_

# Metadata association

## Set up

Please provide the URL to your google sheet below. 

A sample URL has been provided for a google sheet containing MAC season 8 metadata. Replace the sample URL with your own.

```{r url}

url <- 'https://docs.google.com/spreadsheets/d/1BPkZPfynZI9DB9bRJfa335uC_uNjST5jHVf-XUgSVvM/edit#gid=347205052' 

```

The following packages will need to be loaded: RPostgres, readxl, dplyr, kableExtra, googlesheets and glue

```{r load-pack, message = FALSE}

library(RPostgres)
library(readxl)
library(dplyr)
library(kableExtra)
library(googlesheets)
library(glue)

```

A connection to the test database needs to be created. Make sure that you have the test database running and your new metadata has been added (from part 1).

The port number should be the same as the one indicated in docker-compose.override.yml.

```{r create-dbcon}

dbcon <- dbConnect(RPostgres::Postgres(),
                   host = 'localhost',
                   user = 'bety',
                   port = 5433)

```

```{r pack-fx, echo = FALSE}
####################
# exported functions
####################

# add_experiments
add_experiments <- function(experiments, user_id, dbcon, commit){
  
  # for each row of experiments, upload data using add_row
  apply(experiments,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'experiments',
        tbl_fields = c('name', 'start_date', 'end_date',
                       'description', 'design'),
        user_id = user_id,
        commit = commit)
}

# add_sites
add_sites <- function(sites, user_id, dbcon, commit){
  # for each row of sites, upload data using add_row
  apply(sites,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'sites',
        tbl_fields = c('city', 'state', 'country',
                       'notes', 'sitename', 'greenhouse',
                       'geometry', 'time_zone'),
        user_id = user_id,
        commit = commit)
}

# add_treatments
add_treatments <- function(treatments, user_id, dbcon, commit){
  # for each row of treatments, upload data using add_row
  apply(treatments,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'treatments',
        tbl_fields = c('name', 'definition', 'control'),
        user_id = user_id,
        commit = commit)
}

# add_species
add_species <- function(cultivars, dbcon){
  # need to determine which species to add
  bety_species <- tbl(dbcon, 'species') %>% collect()
  unq_spp <- unique(cultivars$species)
  spp_to_add <- unq_spp[!unq_spp %in% bety_species$scientificname]
  if(length(spp_to_add) != 0){
    for(spp in spp_to_add){
      spp_insert <- glue::glue_sql("insert into species (scientificname) values ({spp})",
                                   .con = dbcon)
      send <- DBI::dbSendStatement(dbcon, spp_insert)
      DBI::dbClearResult(send)
    }
  }
}

# get_new_cultivars 
get_new_cultivars <- function(cultivars, dbcon){
  # create specie_id column in cultivars 
  bety_species <- tbl(dbcon, 'species') %>% collect() # read in updated bety species table
  cultivars$specie_id <- vector('numeric', nrow(cultivars)) # create specie_id column
  for(i in 1:nrow(cultivars)){ # get specie id from specie name
    sci_name <- cultivars$species[i]
    spp_id <- bety_species %>%
      filter(scientificname == sci_name) %>%
      select(id)
    cultivars$specie_id[i] <- as.double(spp_id$id)
  }
  # get subset of cultivars to be uploaded
  bety_cultivars <- tbl(dbcon, 'cultivars') %>% 
    select(name, specie_id) %>%
    collect()
  bety_cultivars$specie_id <- as.double(bety_cultivars$specie_id)
  new_cultivars <- anti_join(cultivars,
                             bety_cultivars,
                             by = c('name', 'specie_id'))
  return(new_cultivars)
}

# add_cultivars
add_cultivars <- function(new_cultivars, dbcon, commit){
  # for each row of new cultivars, upload data using add_row
  apply(new_cultivars,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'cultivars',
        tbl_fields = c('name', 'specie_id', 'ecotype', 'notes'),
        commit = commit)
}

# add_citations
add_citations <- function(citations, user_id, dbcon, commit){
  # for each row of citations, upload data using add_row
  apply(citations,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'citations',
        tbl_fields = c('author', 'year', 'title',
                       'journal', 'volume', 'page',
                       'url', 'pdf', 'doi'),
        user_id = user_id,
        commit = commit)
}

# associate_experiments_sites
associate_experiments_sites <- function(sites, dbcon, commit){
  bety_sites <- tbl(dbcon, 'sites') %>% collect() # bety sites table
  bety_experiments <- tbl(dbcon, 'experiments') %>% collect() # bety experiments table
  new_site <- sites$sitename # new sites added
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    exp_id <- bety_experiments %>%
      filter(name == row$experiment) %>%
      select(id)
    associate_vals(tbl_name = 'experiments_sites',
                   col_1 = 'experiment_id',
                   col_2 = 'site_id',
                   val_1 = as.double(exp_id$id),
                   val_2 = as.double(site_id$id),
                   dbcon = dbcon,
                   commit = commit)
  }
}

# associate_experiments_treatments
associate_experiments_treatments <- function(treatments, dbcon, commit){
  bety_treatments <- tbl(dbcon, 'treatments') %>% collect() # bety treatments table
  bety_experiments <- tbl(dbcon, 'experiments') %>% collect() # bety experiments table
  new_treat <- treatments$name # new treatments
  for(treat in new_treat){
    row <- treatments %>% filter(name == treat)
    treat_id <- bety_treatments %>%
      filter(name == treat) %>%
      select(id)
    associate_exp <- strsplit(row$experiment, split = ', ')[[1]]
    for(exp in associate_exp){
      exp_id <- bety_experiments %>%
        filter(name == exp) %>%
        select(id)
      associate_vals(tbl_name = 'experiments_treatments',
                     col_1 = 'experiment_id',
                     col_2 = 'treatment_id',
                     val_1 = as.double(exp_id$id),
                     val_2 = as.double(treat_id$id), 
                     dbcon = dbcon,
                     commit = commit)
    }
  }
}

# associate_sites_cultivars
associate_sites_cultivars <- function(sites, dbcon, commit){
  bety_sites <- tbl(dbcon, 'sites') %>% collect() # bety sites table
  bety_species <- tbl(dbcon, 'species') %>% collect() # bety species table
  bety_cultivars <- tbl(dbcon, 'cultivars') %>% collect() # bety cultivars table
  new_site <- sites$sitename
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    spp_id <- bety_species %>%
      filter(scientificname == row$species) %>%
      select(id)
    cultivar_id <- bety_cultivars %>%
      filter(name == row$cultivar & specie_id == as.double(spp_id$id)) %>%
      select(id)
    associate_vals(tbl_name = 'sites_cultivars',
                   col_1 = 'site_id',
                   col_2 = 'cultivar_id',
                   val_1 = as.double(site_id$id),
                   val_2 = as.double(cultivar_id$id),
                   dbcon = dbcon,
                   commit = commit)
  }
}

#associate_citations_sites
associate_citations_sites <- function(citations, sites, dbcon, commit){
  bety_citations <- tbl(dbcon, 'citations') %>% collect()
  bety_sites <- tbl(dbcon, 'sites') %>% collect()
  # get citation id (assuming only one citation)
  citation_id <- bety_citations %>%
    filter(author == citations$author & year == citations$year & title == citations$title) %>%
    select(id)
  new_site <- sites$sitename # new sites added
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    associate_vals(tbl_name = 'citations_sites',
                   col_1 = 'citation_id',
                   col_2 = 'site_id',
                   val_1 = as.double(citation_id$id),
                   val_2 = as.double(site_id$id), 
                   dbcon = dbcon, 
                   commit = commit)
  }
}

###################
# helper functions
###################

prepared_statement <- function(con, query, params) {
  stopifnot(
    class(con) == "PqConnection",
    is.character(query),
    length(query) == 1,
    is.list(params)
  )
  qry <- DBI::dbSendStatement(con, query)
  res <- DBI::dbBind(qry, params)
  on.exit(DBI::dbClearResult(res))
}

get_fields <- function(tbl_fields, row){
  com_ind <- which(!is.na(row))
  com_fields <- names(row)[com_ind]
  com_tbl_fields <- tbl_fields[tbl_fields %in% com_fields]
  return(com_tbl_fields)
}

get_params <- function(row, fields){
  params <- unname(as.list(row[fields]))
  return(params)
}

get_statement <- function(dbcon, tbl_name, fields){
  num_params <- length(fields)
  posit_params <- sapply(1:num_params, function(x) paste0('$', x))
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)}
                              ({DBI::SQL(paste(fields, collapse = ', '))})
                              values ({DBI::SQL(paste(posit_params, collapse = ', '))})",
                              .con = dbcon)
  return(statement)
}

add_row <- function(row, dbcon, tbl_name, tbl_fields, user_id = NULL, commit = FALSE){
  fields <- get_fields(tbl_fields, row)
  params <- get_params(row, fields)
  if(tbl_name != 'cultivars'){ # cultivars does not have user id column # has a previous id column
    fields <- append(fields, 'user_id')
    params <- append(params, as.double(user_id))
  }
  statement <- get_statement(dbcon, tbl_name, fields)
  DBI::dbBegin(dbcon)
  prepared_statement(dbcon, statement, params)
  ifelse(commit == TRUE, DBI::dbCommit(dbcon), DBI::dbRollback(dbcon))
}

associate_vals <- function(tbl_name, col_1, col_2, val_1, val_2, dbcon, commit = FALSE){
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)}
                              ({DBI::SQL(paste(c(col_1, col_2), collapse = ', '))})
                              values ($1, $2)",
                              .con = dbcon)
  params <- list(val_1, val_2)
  DBI::dbBegin(dbcon)
  prepared_statement(dbcon, statement, params)
  ifelse(commit == TRUE, DBI::dbCommit(dbcon), DBI::dbRollback(dbcon))
}

```

Read in the users, sites, treatments, and citations sheets. 

```{r read-user-gs, message = FALSE}

# get key from URL
key <- extract_key_from_url(url)

# register sheet 
gs_obj <- key %>% gs_key(lookup = FALSE)

# users sheet
users <- gs_obj %>% gs_read(ws = 'users')

# sites sheet
sites <- gs_obj %>% gs_read(ws = 'sites')

# treatments sheet
treatments <- gs_obj %>% gs_read(ws = 'treatments')

# citations sheet
citations <- gs_obj %>% gs_read(ws = 'citations')

```

## Associate data

The following functions will be used to associate metadata in BETYdb: `associate_experiments_sites`, `associate_experiments_treatments`, `associate_sites_cultivars`, and `associate_citations_sites`. These functions are from _packagename_.

### Step 1: Associate experiments with sites

To associate experiments with sites, run `associate_experiments_sites`. The sites google sheet object will need to be passed in. 

```{r associate-exp-site}

associate_experiments_sites(sites, dbcon, commit)

```

### Step 2: Associate experiments with treatments

To associate experiments with treatments, run `associate_experiments_treatments`. The treatments google sheet object will need to be passed in.

```{r associate-exp-treat}

associate_experiments_treatments(treatments, dbcon, commit)

```

### Step 3: Associate sites with cultivars

To associate sites with cultivars, run `associate_sites_cultivars`. The sites google sheet object will need to be passed in.

```{r associate-site-cultivar}

associate_sites_cultivars(sites, dbcon, commit)

```

### Step 4: Associate citations with sites

To associate citations with sites, run `associate_citations_sites`. The citations and sites google sheet objects will need to be passed in.

```{r associate-citation-site}

associate_citations_sites(citations, sites, dbcon, commit)

```

## Check data association

You can check if your data has been correctly associated in the database once you run this markdown file with `commit <- TRUE` on line 26.

These chunks will only be evaluated if `commit <- TRUE`. If evaluated, you should see a table of fetched records below a chunk.

We will be using your user id and today's date to fetch the data you uploaded.

```{r today}

today <- Sys.Date()

# get user id from user login
bety_users <- tbl(dbcon, 'users') %>% collect()
  
user_id <- bety_users %>%
  filter(login == users$login) %>%
  select(id)

```

### Fetch records from experiments_sites

```{r fetch-exp-site, eval = eval_chk}

exp_site_query <- dbSendQuery(dbcon,
                              glue_sql("select * from experiments_sites where experiment_id in 
                                       (select id from experiments where created_at::date = {today} and 
                                       user_id = (select id from users where login = {users$login})) and 
                                       site_id in (select id from sites where created_at::date = {today} and
                                       user_id = (select id from users where login = {users$login})) limit 10",
                                       .con = dbcon))

exp_site_fetch <- dbFetch(exp_site_query)
dbClearResult(exp_site_query)

kable(exp_site_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

## Fetch records from experiments_treatments

```{r fetch-exp-treat, eval = eval_chk}

exp_treat_query <- dbSendQuery(dbcon,
                               glue_sql("select * from experiments_treatments where experiment_id in
                                        (select id from experiments where created_at::date = {today} and
                                        user_id = (select id from users where login = {users$login})) and                                                  treatment_id in (select id from treatments where created_at::date = {today}                                         and user_id = (select id from users where login = {users$login}))",
                                        .con = dbcon))

exp_treat_fetch <- dbFetch(exp_treat_query)
dbClearResult(exp_treat_query)

kable(exp_treat_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

### Fetch records from sites_cultivars

```{r fetch-site-cultivar, eval = eval_chk}

site_cultivar_query <- dbSendQuery(dbcon,
                                   glue_sql("select * from sites_cultivars where site_id in
                                            (select id from sites where created_at::date = {today} and 
                                            user_id = (select id from users where login = {users$login})) 
                                            limit 10",
                                            .con = dbcon))

site_cultivar_fetch <- dbFetch(site_cultivar_query)
dbClearResult(site_cultivar_query)

kable(site_cultivar_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

### Fetch records from citations_sites

```{r fetch-citation-site, eval = eval_chk}

citation_site_query <- dbSendQuery(dbcon,
                                   glue_sql("select * from citations_sites where citation_id in
                                            (select id from citations where created_at::date = {today} and 
                                            user_id = (select id from users where login = {users$login})) and 
                                            site_id in (select id from sites where created_at::date = {today} and
                                            user_id = (select id from users where login = {users$login})) 
                                            limit 10",
                                            .con = dbcon))

citation_site_fetch <- dbFetch(citation_site_query)
dbClearResult(citation_site_query)

kable(citation_site_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

## Close database connection

```{r close-dbcon}

dbDisconnect(dbcon)

```
