---
title: 'Metadata Upload Tutorial Part 2: Associations'
author: "Kimberly Huynh"
date: "4/30/2019"
output: html_document
---

# Section 1: Overview

The objective of this tutorial is to demonstrate how to associate metadata that you have added to the traits database. 

This tutorial should be completed after `Metadata Upload Tutorial Part 1: Additions`.

For this tutorial, make sure that you have a local instance of a test database running and that all new experiments, sites, treatments, cultivars, and citations have been added.

## Notes on Running this Rmd

Knitting this markdown file will not associate any data in the test database. Being able to knit this file means that you will not encounter any errors when associating experiments with sites, experiments with treatments, sites with cultivars, and citations with sites.

To make associations in the test database, you will need to make an edit to this markdown file. 

`commit = TRUE` needs to be set in the following chunk.

```{r commit-param}

commit <- TRUE

```

# Section 2: Required input

URL to a public published google sheet (please provide in section 3)

_this should be the same URL that you used for the `Metadata Upload Tutorial Part 1: Addition` tutorial_

# Section 3: Metadata association

## Set up

Please provide the URL to your google sheet below. 

A sample URL has been provided for a google sheet containing MAC season 8 metadata. Replace the sample URL with your own.

```{r url}

url <- 'https://docs.google.com/spreadsheets/d/1BPkZPfynZI9DB9bRJfa335uC_uNjST5jHVf-XUgSVvM/edit#gid=0' 

```

The following packages will need to be loaded: RPostgres, readxl, dplyr, and googlesheets

```{r load-pack, message = FALSE}

library(RPostgres)
library(readxl)
library(dplyr)
library(googlesheets)

```

Connect to database. Make sure that you have the test database running and your new metadata has been added.

The port number should be the same as the one indicated in docker-compose.override.yml.

```{r create-dbcon}

dbcon <- dbConnect(RPostgres::Postgres(),
                   host = 'localhost',
                   user = 'bety',
                   port = 5433)

```

The following functions will be included in the package.

```{r pack-fx}

# add_experiments
add_experiments <- function(experiments, users, dbcon, commit = FALSE){
  
  # get user id from user login
  bety_users <- tbl(dbcon, 'users') %>% collect()
  user_id <- bety_users %>%
    filter(login == users$login) %>%
    select(id)
  
  # for each row of experiments, upload data using add_row
  apply(experiments,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'experiments',
        tbl_fields = c('name', 'start_date', 'end_date',
                       'description', 'design'),
        user_id = user_id$id,
        commit = commit)
}

# add_sites
add_sites <- function(sites, dbcon, commit = FALSE){
  # for each row of sites, upload data using add_row
  apply(sites,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'sites',
        tbl_fields = c('city', 'state', 'country',
                       'notes', 'sitename', 'greenhouse',
                       'geometry', 'time_zone'),
        commit = commit)
}

# add_treatments
add_treatments <- function(treatments, dbcon, commit = FALSE){
  # for each row of treatments, upload data using add_row
  apply(treatments,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'treatments',
        tbl_fields = c('name', 'defintion', 'control'),
        commit = commit)
}

# add_species
add_species <- function(cultivars, dbcon){
  # need to determine which species to add
  bety_species <- tbl(dbcon, 'species') %>% collect()
  unq_spp <- unique(cultivars$species)
  spp_to_add <- unq_spp[!unq_spp %in% bety_species$scientificname]
  if(length(spp_to_add) != 0){
    for(spp in spp_to_add){
      spp_insert <- glue::glue_sql("insert into species (scientificname) values ({spp})",
                                   .con = dbcon)
      send <- DBI::dbSendStatement(dbcon, spp_insert)
      DBI::dbClearResult(send)
    }
  }
}

# get_new_cultivars 
get_new_cultivars <- function(cultivars, dbcon){
  # create specie_id column in cultivars 
  bety_species <- tbl(dbcon, 'species') %>% collect() # read in updated bety species table
  cultivars$specie_id <- vector('numeric', nrow(cultivars)) # create specie_id column
  for(i in 1:nrow(cultivars)){ # get specie id from specie name
    sci_name <- cultivars$species[i]
    spp_id <- bety_species %>%
      filter(scientificname == sci_name) %>%
      select(id)
    cultivars$specie_id[i] <- as.double(spp_id$id)
  }
  # get subset of cultivars to be uploaded
  bety_cultivars <- tbl(dbcon, 'cultivars') %>% 
    select(name, specie_id) %>%
    collect()
  bety_cultivars$specie_id <- as.double(bety_cultivars$specie_id)
  new_cultivars <- anti_join(cultivars[, c('name', 'specie_id')],
                             bety_cultivars,
                             by = c('name', 'specie_id'))
  return(new_cultivars)
}

# add_cultivars
add_cultivars <- function(new_cultivars, dbcon, commit = FALSE){
  # for each row of new cultivars, upload data using add_row
  apply(new_cultivars,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'cultivars',
        tbl_fields = c('name', 'specie_id', 'ecotype', 'notes'),
        commit = commit)
}

# add_citations
add_citations <- function(citations, dbcon, commit = FALSE){
  # for each row of citations, upload data using add_row
  apply(citations,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'citations',
        tbl_fields = c('author', 'year', 'title',
                       'journal', 'volume', 'page',
                       'url', 'pdf', 'doi'),
        commit = commit)
}

# associate_experiments_sites
associate_experiments_sites <- function(sites, dbcon, commit){
  bety_sites <- tbl(dbcon, 'sites') %>% collect() # bety sites table
  bety_experiments <- tbl(dbcon, 'experiments') %>% collect() # bety experiments table
  new_site <- sites$sitename # new sites added
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    exp_id <- bety_experiments %>%
      filter(name == row$experiment) %>%
      select(id)
    associate_vals(tbl_name = 'experiments_sites',
                   col_1 = 'experiment_id',
                   col_2 = 'site_id',
                   val_1 = as.double(exp_id$id),
                   val_2 = as.double(site_id$id),
                   dbcon = dbcon,
                   commit = commit)
  }
}

# associate_experiments_treatments
associate_experiments_treatments <- function(treatments, dbcon, commit){
  bety_treatments <- tbl(dbcon, 'treatments') %>% collect() # bety treatments table
  bety_experiments <- tbl(dbcon, 'experiments') %>% collect() # bety experiments table
  new_treat <- treatments$name # new treatments
  for(treat in new_treat){
    row <- treatments %>% filter(name == treat)
    treat_id <- bety_treatments %>%
      filter(name == treat) %>%
      select(id)
    associate_exp <- strsplit(row$experiment, split = ', ')[[1]]
    for(exp in associate_exp){
      exp_id <- bety_experiments %>%
        filter(name == exp) %>%
        select(id)
      associate_vals(tbl_name = 'experiments_treatments',
                     col_1 = 'experiment_id',
                     col_2 = 'treatment_id',
                     val_1 = as.double(exp_id$id),
                     val_2 = as.double(treat_id$id), 
                     dbcon = dbcon,
                     commit = commit)
    }
  }
}

# associate_sites_cultivars
associate_sites_cultivars <- function(sites, dbcon, commit){
  bety_sites <- tbl(dbcon, 'sites') %>% collect() # bety sites table
  bety_species <- tbl(dbcon, 'species') %>% collect() # bety species table
  bety_cultivars <- tbl(dbcon, 'cultivars') %>% collect() # bety cultivars table
  new_site <- sites$sitename
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    spp_id <- bety_species %>%
      filter(scientificname == row$species) %>%
      select(id)
    cultivar_id <- bety_cultivars %>%
      filter(name == row$cultivar & specie_id == as.double(spp_id$id)) %>%
      select(id)
    associate_vals(tbl_name = 'sites_cultivars',
                   col_1 = 'site_id',
                   col_2 = 'cultivar_id',
                   val_1 = as.double(site_id$id),
                   val_2 = as.double(cultivar_id$id),
                   dbcon = dbcon,
                   commit = commit)
  }
}

#associate_citations_sites
associate_citations_sites <- function(citations, sites, dbcon, commit){
  bety_citations <- tbl(dbcon, 'citations') %>% collect()
  bety_sites <- tbl(dbcon, 'sites') %>% collect()
  # get citation id (assuming only one citation)
  citation_id <- bety_citations %>%
    filter(author == citations$author & year == citations$year & title == citations$title) %>%
    select(id)
  new_site <- sites$sitename # new sites added
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    associate_vals(tbl_name = 'citations_sites',
                   col_1 = 'citation_id',
                   col_2 = 'site_id',
                   val_1 = as.double(citation_id$id),
                   val_2 = as.double(site_id$id), 
                   dbcon = dbcon, 
                   commit = commit)
  }
}

###################
# helper functions
###################

prepared_statement <- function(con, query, params) {
  stopifnot(
    class(con) == "PqConnection",
    is.character(query),
    length(query) == 1,
    is.list(params)
  )
  qry <- DBI::dbSendStatement(con, query)
  res <- DBI::dbBind(qry, params)
  on.exit(DBI::dbClearResult(res))
}

get_fields <- function(tbl_fields, row){
  com_ind <- which(!is.na(row))
  com_fields <- names(row)[com_ind]
  com_tbl_fields <- tbl_fields[tbl_fields %in% com_fields]
  return(com_tbl_fields)
}

get_params <- function(row, fields){
  params <- unname(as.list(row[fields]))
  return(params)
}

get_statement <- function(dbcon, tbl_name, fields){
  num_params <- length(fields)
  posit_params <- sapply(1:num_params, function(x) paste0('$', x))
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)}
                              ({DBI::SQL(paste(fields, collapse = ', '))})
                              values ({DBI::SQL(paste(posit_params, collapse = ', '))})",
                              .con = dbcon)
  return(statement)
}

add_row <- function(row, dbcon, tbl_name, tbl_fields, user_id = NULL, commit = FALSE){
  #user_id only required for experiment upload

  fields <- get_fields(tbl_fields, row)
  params <- get_params(row, fields)
  # extra addition needed to insert data into experiments
  if(tbl_name == 'experiments'){
    fields <- append(fields, 'user_id')
    params <- append(params, as.double(user_id)) #need to create this in experiments chunk
  }
  statement <- get_statement(dbcon, tbl_name, fields)
  DBI::dbBegin(dbcon)
  prepared_statement(dbcon, statement, params)
  ifelse(commit == TRUE, DBI::dbCommit(dbcon), DBI::dbRollback(dbcon))
}

associate_vals <- function(tbl_name, col_1, col_2, val_1, val_2, dbcon, commit = FALSE){
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)}
                              ({DBI::SQL(paste(c(col_1, col_2), collapse = ', '))})
                              values ($1, $2)",
                              .con = dbcon)
  params <- list(val_1, val_2)
  DBI::dbBegin(dbcon)
  prepared_statement(dbcon, statement, params)
  ifelse(commit == TRUE, DBI::dbCommit(dbcon), DBI::dbRollback(dbcon))
}

```

Read in the treatments, sites, and citations sheets. 

These sheets contain information about which metadata to associate.

```{r read-user-gs, message = FALSE}

# get key from URL
key <- extract_key_from_url(url)

# register sheet 
gs_obj <- key %>% gs_key(lookup = FALSE)

# sites sheet
sites <- gs_obj %>% gs_read(ws = 'sites')

# treatments sheet
treatments <- gs_obj %>% gs_read(ws = 'treatments')

# citations sheet
citations <- gs_obj %>% gs_read(ws = 'citations')

```

## Associate data

### Step 1: Associate experiments with sites

To associate experiments with sites, run `associate_experiments_sites`. The sites google sheet object will need to be passed in. 

```{r associate-exp-site}

associate_experiments_sites(sites, dbcon, commit)

```

### Step 2: Associate experiments with treatments

To associate experiments with treatments, run `associate_experiments_treatments`. The treatments google sheet object will need to be passed in.

```{r associate-exp-treat}

associate_experiments_treatments(treatments, dbcon, commit)

```

### Step 3: Associate sites with cultivars

To associate sites with cultivars, run `associate_sites_cultivars`. The sites google sheet object will need to be passed in.

```{r associate-site-cultivar}

associate_sites_cultivars(sites, dbcon, commit)

```

### Step 4: Associate citations with sites

To associate citations with sites, run `associate_citations_sites`. The citations and sites google sheet objects will need to be passed in.

```{r associate-citation-site}

associate_citations_sites(citations, sites, dbcon, commit)

```

## Close database connection

```{r close-dbcon}

dbDisconnect(dbcon)

```
