---
title: "Metadata Upload Tutorial"
author: "Kimberly Huynh"
date: "4/23/2019"
output: html_document
---

# Section 1: Overview

The objective of this tutorial is to demonstrate how to upload a season's metadata to the traits database.

Each season should have metadata for experiments, sites, treatments, cultivars, and citations.

For this tutorial, data will be uploaded to a test database. Instructions on how to set up a local instance of a test database can be found [here](https://gist.github.com/kimberlyh66/935106e35478829b9e844ba5b496784d).

**need to update location of setting up local instance of test db tutorial**

# Section 2: Required input

URL to a public published google sheet

   * copy and fill out the following [google sheets template](https://docs.google.com/spreadsheets/d/1Fr_8xYOucyCQ9WH5_1bfPTMpm3IhKX5oqW4UOMvPoSY/edit?usp=sharing).
   * instructions on how to fill out the template can be found under the README worksheet.
   * make the google sheet public (to anyone with link) and publish to the web.
   

Please provide the URL to your google sheet below. 

A sample URL has been provided for a google sheet containing MAC season 8 metadata. Please replace the sample URL with your own.

```{r url}

url <- 'https://docs.google.com/spreadsheets/d/1c_5j7q3TO6gQ24KSopE-_LXA5HhfsUEqMupckGZAbo8/edit#gid=0' 

```

# Section 3: Metadata upload

## Set up

Load in necessary packages

```{r load-pack}

library(RPostgres)
library(readxl)
library(dplyr)
library(googlesheets)

```

Create database connection 

```{r create-dbcon}

# no password required 
dbcon <- dbConnect(RPostgres::Postgres(),
                   host = 'localhost',
                   user = 'bety',
                   port = 5433)

```

Functions to include in package

```{r pack-fx}

# preparedStatement
# uploads parameterized query to database
preparedStatement <- function(con, query, params) {
  stopifnot(
    class(con) == "PqConnection",
    is.character(query),
    length(query) == 1,
    is.list(params)
  )
  qry <- DBI::dbSendStatement(con, query)
  res <- DBI::dbBind(qry, params)
  on.exit(DBI::dbClearResult(res))
}

# getFields
# returns column names of row that have a value and is included in tbl_fields
getFields <- function(tbl_fields,
                      row){
  com_ind <- which(!is.na(row))
  com_fields <- names(row)[com_ind]
  com_tbl_fields <- tbl_fields[tbl_fields %in% com_fields]
  return(com_tbl_fields)
}

# getParams
# returns an unnamed list of parameter values for a row that is subsetted according to fields
getParams <- function(row,
                      fields){
  params <- unname(as.list(row[fields]))
  return(params)
}

# getPositParams
# returns something like '$1, $2, $3' to be used for parameterized queries; number of params depends on length of fields
getPositParams <- function(fields){
  posit_params <- sapply(1:length(fields),
                         function(x) paste0('$', x))
  return(posit_params)
}

# getStatement
# returns a parameterized insert statement 
getStatement <- function(dbcon,
                         tbl_name,
                         fields){
  posit_params <- sapply(1:length(fields),
                         function(x) paste0('$', x))
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)} 
                              ({DBI::SQL(paste(fields, collapse = ', '))}) 
                              values 
                              ({DBI::SQL(paste(posit_params, collapse = ', '))})",
                              .con = dbcon)
  return(statement)
}

# addRow
# uploads insert statement 
# to be used to add new experiments, sites, treatments, cultivars, and citations
addRow <- function(row,
                   dbcon,
                   tbl_name,
                   tbl_fields,
                   user_id = NULL){ #user_id only required for experiment upload
  fields <- getFields(tbl_fields,
                      row)
  params <- getParams(row,
                      fields)
  # extra addition needed to insert data into experiments 
  if(tbl_name == 'experiments'){
    fields <- append(fields, 'user_id')
    params <- append(params, as.double(user_id)) #need to create this in experiments chunk
  }
  statement <- getStatement(dbcon,
                            tbl_name,
                            fields)
  preparedStatement(dbcon,
                    statement,
                    params)
}

# associateRow
# uploads insert statement
# to be used to associate experiments with sites, experiments with treatments, sites with cultivars, and sites with citations
associateRow <- function(tbl_name,
                         col_1,
                         col_2,
                         val_1,
                         val_2){
    statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)} 
                                ({DBI::SQL(paste(c(col_1, col_2), collapse = ', '))}) 
                                values ($1, $2)",
                                .con = dbcon)
    params <- list(val_1, val_2)
    preparedStatement(dbcon,
                      statement,
                      params)
}

```

Read in metadata from google sheets

```{r read-user-gs}

# get key from URL
key <- extract_key_from_url(url)

# register sheet 
gs_obj <- key %>% gs_key(lookup = FALSE)

# users sheet
users <- gs_obj %>% gs_read(ws = 'users')

# experiments sheet
experiments <- gs_obj %>% gs_read(ws = 'experiments')

# sites sheet
sites <- gs_obj %>% gs_read(ws = 'sites')

# treatments sheet
treatments <- gs_obj %>% gs_read(ws = 'treatments')

# cultivars sheet
cultivars <- gs_obj %>% gs_read(ws = 'cultivars')

# citations sheet
citations <- gs_obj %>% gs_read(ws = 'citations')

```


## Upload data

### Addition steps

#### Step 1: Add new experiments

```{r add-exp}

# need user id to insert data into experiments table
# refer to bety users table to get user id from user login (provided by user)
bety_users <- tbl(dbcon, 'users') %>% collect()
 
user_id <- bety_users %>%
  filter(login == users$login) %>%
  select(id)

# get names of new experiments
new_exp <- experiments$name # assuming that each row of experiments sheet describes a unique experiment

# loop through new experiments
# filter for row with experiment name
# upload data for row using addRow
for(exp in new_exp){
  row <- experiments %>% filter(name == exp)
  addRow(row = row,
         dbcon = dbcon,
         tbl_name = 'experiments',
         tbl_fields = c('name', 'start_date', 'end_date',
                        'description', 'design'),
         user_id = user_id$id) #need to specify user_id since adding experiment
}

```

#### Step 2: Add new sites

```{r add-site}

# get new site names
new_site <- sites$sitename # assuming that each row of sites sheet describes a unique site

# loop through new sitenames
# filter for row with sitename
# upload data for row using addRow
for(site in new_site){
  row <- sites %>% filter(sitename == site)
  addRow(row = row,
         dbcon = dbcon,
         tbl_name = 'sites',
         tbl_fields = c('city', 'state', 'country',
                        'notes', 'sitename', 'greenhouse',
                        'geometry', 'time_zone'))
}

```

#### Step 3: Add new treatments

```{r add-treat}

# get new treatments
new_treat <- treatments$name # assuming that each row of treatments sheet describes a unique treatment

# loop through new treatments
# filter for row with treatment name
# upload data for row using addRow
for(treat in new_treat){
  row <- treatments %>% filter(name == treat)
  addRow(row = row,
         dbcon = dbcon,
         tbl_name = 'treatments',
         tbl_fields = c('name', 'defintion', 'control'))
}

```

#### Step 4: Add new cultivars

Only rows that contain a combination of cultivar name and specie id that is not yet present in BETYdb will be uploaded.

First, must make sure all species present in cultivars sheet have been added to BETY species table.

```{r check-spp}

# refer to bety species table
bety_species <- tbl(dbcon, 'species') %>% collect()

# get unique species in cultivars
unq_spp <- unique(cultivars$species)

# get species to add
spp_to_add <- unq_spp[!unq_spp %in% bety_species$scientificname]

if(length(spp_to_add) != 0){
  for(spp in spp_to_add){
    spp_insert <- glue::glue_sql("insert into species (scientificname) values ({spp})",
                                 .con = dbcon)
  }
}

```

Now, create specie_id column in cultivars.

```{r add-spp-id}

cultivars$specie_id <- vector('numeric', nrow(cultivars))

# get specie id from specie name
for(i in 1:nrow(cultivars)){
  sci_name <- cultivars$species[i]
  spp_id <- bety_species %>%
    filter(scientificname == sci_name) %>%
    select(id)
  cultivars$specie_id[i] <- as.double(spp_id$id)
}
  
```

Determine which combinations of cultivar name and species id should be uploaded to bety

```{r get-new-cultivars}

# refer to bety cultivars table
bety_cultivars <- tbl(dbcon, 'cultivars') %>% 
  select(name, specie_id) %>%
  collect()

# want data type of specie_id columns in cultivars and bety_cultivars to be the same
bety_cultivars$specie_id <- as.double(bety_cultivars$specie_id)

# get subset of new cultivar + specie_id combinations
new_cultivars <- anti_join(cultivars[, c('name', 'specie_id')],   
                           bety_cultivars,
                           by = c('name', 'specie_id'))

```

Add new cultivars

```{r add-cultivar}

# loop through each row of new_cultivars
# upload data for row using addRow
for(i in 1:nrow(new_cultivars)){
  row <- new_cultivars %>% slice(i)
  addRow(row = row,
         dbcon = dbcon,
         tbl_name = 'cultivars',
         tbl_fields = c('name', 'specie_id', 'ecotype', 'notes'))
}

```

#### Step 5: Add new citation

```{r add-citation}

# loop through each row of citations
# upload data for row using addRow

for(i in 1:nrow(citations)){
  row <- citations %>% slice(i)
  addRow(row = row,
         dbcon = dbcon,
         tbl_name = 'citations',
         tbl_fields = c('author', 'year', 'title',
                        'journal', 'volume', 'page',
                        'url', 'pdf', 'doi'))
}

```

#### Association steps

Read in updated bety experiments, sites, cultivars, treatments, and citations tables after new additions have been made

```{r read-bety-update}

bety_experiments <- tbl(dbcon, 'experiments') %>% collect()
bety_sites <- tbl(dbcon, 'sites') %>% collect()
bety_cultivars <- tbl(dbcon, 'cultivars') %>% collect()
bety_treatments <- tbl(dbcon, 'treatments') %>% collect()
bety_citations <- tbl(dbcon, 'citations') %>% collect()
bety_species <- tbl(dbcon, 'species') %>% collect()

```

#### Step 6: Associate experiments with sites

```{r associate-exp-site}

# loop through new sites added
# filter for row with sitename
# get site id from sitename
# get experiment id from associated experiment name

for(site in new_site){
  row <- sites %>% filter(sitename == site)
  
  site_id <- bety_sites %>%
    filter(sitename == row$sitename) %>%
    select(id)
  
  exp_id <- bety_experiments %>%
    filter(name == row$experiment) %>%
    select(id)
  
  associateRow(tbl_name = 'experiments_sites',
               col_1 = 'experiment_id',
               col_2 = 'site_id',
               val_1 = as.double(exp_id$id),
               val_2 = as.double(site_id$id))
}


```

### Step 7: Associate experiments with treatments

```{r associate-exp-treat}

# loop through new treatments added
# filter for row with treatment name
# get treatment id
# get vector of associated experiment name
# loop through vector of exp names
# get experiment id

for(treat in new_treat){
  row <- treatments %>% filter(name == treat)
  
  treat_id <- bety_treat %>%
      filter(name == treat) %>%
      select(id)
  
  associate_exp <- strsplit(row$experiment, split = ', ')[[1]]
  
  for(exp in associate_exp){
    exp_id <- bety_experiments %>%
      filter(name == exp) %>%
      select(id)
    
    associateRow(tbl_name = 'experiments_treatments',
               col_1 = 'experiment_id',
               col_2 = 'treatment_id',
               val_1 = as.double(exp_id$id),
               val_2 = as.double(treat_id$id))
  }
}

```

### Step 8: Associate sites with cultivars

```{r associate-site-cultivar}

# loop through new sites added
# filter for row with sitename
# get site_id
# get associated cultivar_id (need specie_id for this)
for(site in new_site){
  row <- sites %>% filter(sitename == site)
   
  site_id <- bety_sites %>%
    filter(sitename == row$sitename) %>%
    select(id)
  
  spp_id <- bety_species %>%
    filter(scientificname == row$species) %>%
    select(id)
  
  cultivar_id <- bety_cultivars %>%
    filter(name == row$cultivar & specie_id == as.double(spp_id$id)) %>%
    select(id)
  
  associateRow(tbl_name = 'sites_cultivars',
               col_1 = 'site_id',
               col_2 = 'cultivar_id',
               val_1 = as.double(site_id$id),
               val_2 = as.double(cultivar_id$id))
}


```

### Step 9: Associate citations with sites

```{r associate-citation-site}

# get citation id
citation_id <- bety_citations %>%
  filter(author == citations$author & year == citations$year & title == citations$title) %>%
  select(id)

# loop through new sites added
# filter for row with sitename
# get site id
# associate with above citation id

for(site in new_site){
  row <- sites %>% filter(sitename == site)
  
  site_id <- bety_sites %>%
    filter(sitename == row$sitename) %>%
    select(id)
  
  associateRow(tbl_name = 'citations_sites',
               col_1 = 'citation_id',
               col_2 = 'site_id',
               val_1 = as.double(citation_id$id),
               val_2 = as.double(site_id$id))
  
}

```

Close database connection

```{r close-dbcon}

dbDisconnect(dbcon)

```
