---
title: "Metadata Upload Tutorial Part 1: Additions"
author: "Kimberly Huynh"
date: "4/30/2019"
output: html_document
---

# Section 1: Overview

The objective of this tutorial is to demonstrate how to upload a season's metadata to the traits database.

Each season should have metadata for experiments, sites, treatments, cultivars, and citations.

For this tutorial, data will be uploaded to a test database. Instructions on how to set up a local instance of a test database can be found [here](https://gist.github.com/kimberlyh66/935106e35478829b9e844ba5b496784d).

This tutorial will only cover how to add new experiments, sites, treatments, cultivars, and citations. 

The `Metadata Upload Tutorial Part 2: Associations` tutorial should be viewed after completing this tutorial and when all new metadata have been added (see below for more details). The second tutorial will cover how to associate the metadata that you add.

# Section 2: Required input

URL to a public published google sheet (please provide in setup of section 3)

   * copy and fill out the following [google sheets template](https://docs.google.com/spreadsheets/d/1pNEuwNHNHHrpql0b36DYFAbOlBU5TJVF_nO4RdRzdmg/edit#gid=0).
   * instructions on how to fill out the template can be found under the README worksheet.
   * make the google sheet public (to anyone with link) and publish to the web.

# Section 3: How to Run this Rmd

## Commit database transactions to upload data

Knitting this markdown file will not upload any data to the test database. Being able to knit this file means that you will be able to upload all of your new experiments, sites, treatments, cultivars, and citations metadata successfully.

To actually upload your data to the test database, you will need to make an edit to this markdown file. 

`commit = TRUE` needs to be set in calls to `addRow`. The value of `commit` needs to be changed from `FALSE` to `TRUE` in the following chunk:

```{r commit-param}

commit <- FALSE

```

NOTE: You will get errors if you try to insert your metadata more than once. 

If you would like to restart the test database (remove additions that you have made), run step 4 of the `How to set up local instance of BETY test database` tutorial.

## Check uploaded data

You will not be able to fetch any of your uploaded data until it has been added to the database. You can check your uploaded results when you run this Rmd with `commit <- TRUE` in the above chunk:

```{r check-upload}

ifelse(commit == TRUE, eval_chk <- TRUE, eval_chk <- FALSE)

```

# Section 4: Metadata upload

## Set up

Please provide the URL to your google sheet below. 

A sample URL has been provided for a google sheet containing MAC season 8 metadata. Replace the sample URL with your own.

```{r url}

url <- 'https://docs.google.com/spreadsheets/d/1BPkZPfynZI9DB9bRJfa335uC_uNjST5jHVf-XUgSVvM/edit#gid=0' 

```

The following packages will need to be loaded: `RPostgres`, `readxl`, `dplyr`, `googlesheets`, `kableExtra`, `glue`, and `DBI`

```{r load-pack, message = FALSE}

library(RPostgres)
library(readxl)
library(dplyr)
library(googlesheets)
library(kableExtra)
library(glue)
library(DBI)

```

Create database connection (make sure you have test database running). 

The port number should be the same as the one indicated in docker-compose.override.yml.

```{r create-dbcon}

# no password required 
dbcon <- dbConnect(RPostgres::Postgres(),
                   host = 'localhost',
                   user = 'bety',
                   port = 5433)

```

Functions to include in package

```{r pack-fx}
####################
# exported functions
####################

# add_experiments
add_experiments <- function(experiments, users, dbcon, commit = FALSE){
  
  # get user id from user login
  bety_users <- tbl(dbcon, 'users') %>% collect()
  user_id <- bety_users %>%
    filter(login == users$login) %>%
    select(id)
  
  # for each row of experiments, upload data using add_row
  apply(experiments,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'experiments',
        tbl_fields = c('name', 'start_date', 'end_date',
                       'description', 'design'),
        user_id = user_id$id,
        commit = commit)
}

# add_sites
add_sites <- function(sites, dbcon, commit = FALSE){
  # for each row of sites, upload data using add_row
  apply(sites,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'sites',
        tbl_fields = c('city', 'state', 'country',
                       'notes', 'sitename', 'greenhouse',
                       'geometry', 'time_zone'),
        commit = commit)
}

# add_treatments
add_treatments <- function(treatments, dbcon, commit = FALSE){
  # for each row of treatments, upload data using add_row
  apply(treatments,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'treatments',
        tbl_fields = c('name', 'defintion', 'control'),
        commit = commit)
}

# add_species
add_species <- function(cultivars, dbcon){
  # need to determine which species to add
  bety_species <- tbl(dbcon, 'species') %>% collect()
  unq_spp <- unique(cultivars$species)
  spp_to_add <- unq_spp[!unq_spp %in% bety_species$scientificname]
  if(length(spp_to_add) != 0){
    for(spp in spp_to_add){
      spp_insert <- glue::glue_sql("insert into species (scientificname) values ({spp})",
                                   .con = dbcon)
      send <- DBI::dbSendStatement(dbcon, spp_insert)
      DBI::dbClearResult(send)
    }
  }
}

# get_new_cultivars 
get_new_cultivars <- function(cultivars, dbcon){
  # create specie_id column in cultivars 
  bety_species <- tbl(dbcon, 'species') %>% collect() # read in updated bety species table
  cultivars$specie_id <- vector('numeric', nrow(cultivars)) # create specie_id column
  for(i in 1:nrow(cultivars)){ # get specie id from specie name
    sci_name <- cultivars$species[i]
    spp_id <- bety_species %>%
      filter(scientificname == sci_name) %>%
      select(id)
    cultivars$specie_id[i] <- as.double(spp_id$id)
  }
  # get subset of cultivars to be uploaded
  bety_cultivars <- tbl(dbcon, 'cultivars') %>% 
    select(name, specie_id) %>%
    collect()
  bety_cultivars$specie_id <- as.double(bety_cultivars$specie_id)
  new_cultivars <- anti_join(cultivars[, c('name', 'specie_id')],
                             bety_cultivars,
                             by = c('name', 'specie_id'))
  return(new_cultivars)
}

# add_cultivars
add_cultivars <- function(new_cultivars, dbcon, commit = FALSE){
  # for each row of new cultivars, upload data using add_row
  apply(new_cultivars,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'cultivars',
        tbl_fields = c('name', 'specie_id', 'ecotype', 'notes'),
        commit = commit)
}

# add_citations
add_citations <- function(citations, dbcon, commit = FALSE){
  # for each row of citations, upload data using add_row
  apply(citations,
        MARGIN = 1,
        add_row,
        dbcon = dbcon,
        tbl_name = 'citations',
        tbl_fields = c('author', 'year', 'title',
                       'journal', 'volume', 'page',
                       'url', 'pdf', 'doi'),
        commit = commit)
}

# associate_experiments_sites
associate_experiments_sites <- function(sites, dbcon, commit){
  bety_sites <- tbl(dbcon, 'sites') # bety sites table
  bety_experiments <- tbl(dbcon, 'experiments') # bety experiments table
  new_site <- sites$sitename # new sites added
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    exp_id <- bety_experiments %>%
      filter(name == row$experiment) %>%
      select(id)
    associate_vals(tbl_name = 'experiments_sites',
                   col_1 = 'experiment_id',
                   col_2 = 'site_id',
                   val_1 = as.double(exp_id$id),
                   val_2 = as.double(site_id$id),
                   dbcon = dbcon,
                   commit = commit)
  }
}

# associate_experiments_treatments
associate_experiments_treatments <- function(treatments, dbcon, commit){
  bety_treatments <- tbl(dbcon, 'treatments') # bety treatments table
  bety_experiments <- tbl(dbcon, 'experiments') # bety experiments table
  new_treat <- treatments$name # new treatments
  for(treat in new_treat){
    row <- treatments %>% filter(name == treat)
    treat_id <- bety_treatments %>%
      filter(name == treat) %>%
      select(id)
    associate_exp <- strsplit(row$experiment, split = ', ')[[1]]
    for(exp in associate_exp){
      exp_id <- bety_experiments %>%
        filter(name == exp) %>%
        select(id)
      associate_vals(tbl_name = 'experiments_treatments',
                   col_1 = 'experiment_id',
                   col_2 = 'treatment_id',
                   val_1 = as.double(exp_id$id),
                   val_2 = as.double(treat_id$id), 
                   dbcon = dbcon,
                   commit = commit)
    }
  }
}

# associate_sites_cultivars
associate_sites_cultivars <- function(sites, dbcon, commit){
  bety_sites <- tbl(dbcon, 'sites') # bety sites table
  bety_species <- tbl(dbcon, 'species') # bety species table
  bety_cultivars <- tbl(dbcon, 'cultivars') # bety cultivars table
  new_site <- sites$sitename
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    spp_id <- bety_species %>%
      filter(scientificname == row$species) %>%
      select(id)
    cultivar_id <- bety_cultivars %>%
      filter(name == row$cultivar & specie_id == as.double(spp_id$id)) %>%
      select(id)
    associate_vals(tbl_name = 'sites_cultivars',
                   col_1 = 'site_id',
                   col_2 = 'cultivar_id',
                   val_1 = as.double(site_id$id),
                   val_2 = as.double(cultivar_id$id),
                   dbcon = dbcon,
                   commit = commit)
  }
}

#associate_citations_sites
associate_citations_sites <- function(citations, sites, dbcon, commit){
  bety_citations <- tbl(dbcon, 'citations')
  bety_sites <- tbl(dbcon, 'sites')
  # get citation id (assuming only one citation)
  citation_id <- bety_citations %>%
    filter(author == citations$author & year == citations$year & title == citations$title) %>%
    select(id)
  new_site <- sites$sitename # new sites added
  for(site in new_site){
    row <- sites %>% filter(sitename == site)
    site_id <- bety_sites %>%
      filter(sitename == row$sitename) %>%
      select(id)
    associate_vals(tbl_name = 'citations_sites',
                   col_1 = 'citation_id',
                   col_2 = 'site_id',
                   val_1 = as.double(citation_id$id),
                   val_2 = as.double(site_id$id), 
                   dbcon = dbcon, 
                   commit = commit)
  }
}

###################
# helper functions
###################

prepared_statement <- function(con, query, params) {
  stopifnot(
    class(con) == "PqConnection",
    is.character(query),
    length(query) == 1,
    is.list(params)
  )
  qry <- DBI::dbSendStatement(con, query)
  res <- DBI::dbBind(qry, params)
  on.exit(DBI::dbClearResult(res))
}

get_fields <- function(tbl_fields, row){
  com_ind <- which(!is.na(row))
  com_fields <- names(row)[com_ind]
  com_tbl_fields <- tbl_fields[tbl_fields %in% com_fields]
  return(com_tbl_fields)
}

get_params <- function(row, fields){
  params <- unname(as.list(row[fields]))
  return(params)
}

get_statement <- function(dbcon, tbl_name, fields){
  num_params <- length(fields)
  posit_params <- sapply(1:num_params, function(x) paste0('$', x))
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)}
                              ({DBI::SQL(paste(fields, collapse = ', '))})
                              values ({DBI::SQL(paste(posit_params, collapse = ', '))})",
                              .con = dbcon)
  return(statement)
}

add_row <- function(row, dbcon, tbl_name, tbl_fields, user_id = NULL, commit = FALSE){
  #user_id only required for experiment upload

  fields <- get_fields(tbl_fields, row)
  params <- get_params(row, fields)
  # extra addition needed to insert data into experiments
  if(tbl_name == 'experiments'){
    fields <- append(fields, 'user_id')
    params <- append(params, as.double(user_id)) #need to create this in experiments chunk
  }
  statement <- get_statement(dbcon, tbl_name, fields)
  DBI::dbBegin(dbcon)
  prepared_statement(dbcon, statement, params)
  ifelse(commit == TRUE, DBI::dbCommit(dbcon), DBI::dbRollback(dbcon))
}

associate_vals <- function(tbl_name, col_1, col_2, val_1, val_2, dbcon, commit = FALSE){
  statement <- glue::glue_sql("insert into {DBI::SQL(tbl_name)}
                              ({DBI::SQL(paste(c(col_1, col_2), collapse = ', '))})
                              values ($1, $2)",
                              .con = dbcon)
  params <- list(val_1, val_2)
  DBI::dbBegin(dbcon)
  prepared_statement(dbcon, statement, params)
  ifelse(commit == TRUE, DBI::dbCommit(dbcon), DBI::dbRollback(dbcon))
}

```

Read in metadata from google sheets. Each sheet of your google sheet will be read in directly as a R object.

```{r read-user-gs, message = FALSE}

# get key from URL
key <- extract_key_from_url(url)

# register sheet 
gs_obj <- key %>% gs_key(lookup = FALSE)

# users sheet
users <- gs_obj %>% gs_read(ws = 'users')

# experiments sheet
experiments <- gs_obj %>% gs_read(ws = 'experiments')

# sites sheet
sites <- gs_obj %>% gs_read(ws = 'sites')

# treatments sheet
treatments <- gs_obj %>% gs_read(ws = 'treatments')

# cultivars sheet
cultivars <- gs_obj %>% gs_read(ws = 'cultivars')

# citations sheet
citations <- gs_obj %>% gs_read(ws = 'citations')

```


## Upload data

### Step 1: Add new experiments

A user id will be needed to insert data into the experiments table. Your user id will be determined based on your provided user login (username for bety account). 

To add new experiments, run `add_experiments`. The experiments and users google sheet objects will need to be passed in as arguments.

```{r add-experiments}

add_experiments(experiments, users, dbcon, commit)

```

### Step 2: Add new sites

To add new sites, run `add_sites`. The sites google sheet object will need to be passed in.

```{r add-sites}

add_sites(sites, dbcon, commit)

```

### Step 3: Add new treatments

To add new treatments, run `add_treatments`. The treatments google sheet object will need to be passed in.

```{r add-treatments}

add_treatments(treatments, dbcon, commit)

```

### Step 4: Add new cultivars

Only combinations of cultivars and specie id that are not yet present in BETYdb will be uploaded. 

First need to make sure that all species present in cultivars sheet has been added to species table. This is needed since the cultivars table references the species table for specie_id.

If found, new species will be added using `add_species`. The cultivars google sheet object will need to be passed in.

```{r add-species}

add_species(cultivars, dbcon)

```

To determine which cultivars need to be uploaded, run `get_new_cultivars`. The cultivars google spread sheet will need to be passed in.

```{r get-new-cultivars}

new_cultivars <- get_new_cultivars(cultivars, dbcon)

```

Add new cultivars by passing in `new_cultivars` generated in the chunk above to the `add_cultivars` function.

```{r add-cultivars}

add_cultivars(new_cultivars, dbcon, commit)

```

Since this file will be used for testing, new species added by the `get_new_cultivars` function will be removed unless you choose to upload your data. The following chunk will not be run unless you set the value fo `commit` to be `TRUE` on line 40.

```{r remove-spp, eval = eval_chk}

# need to determine if which new species had been added to species table
bety_species <- tbl(dbcon, 'species') %>% collect()

unq_spp <- unique(cultivars$species)

spp_to_add <- unq_spp[!unq_spp %in% bety_species$scientificname]

if(length(spp_to_add) != 0){
  delete_spp <- glue_sql("delete from species where scientificname in
                         ({spp_to_add*})",
                         .con = dbcon)
  send <- dbSendStatement(dbcon, delete_spp)
  dbClearResult(send)
}

```

### Step 5: Add new citations

Add new citations by running `add_citations`. The citations google sheet object will need to be passed in.

```{r add-citations}

add_citations(citations, dbcon, commit)

```

## Check data upload

You can check if your data has been correctly uploaded to the database once you run this markdown file with `commit <- TRUE` on line 40.

These chunks will not be evaluated if `commit <- FALSE`.

### Fetch experiments

```{r fetch-exp, eval = eval_chk}

new_exp <- experiments$name

exp_query <- dbSendQuery(dbcon,
                         glue_sql("select * from experiments where name in ({new_exp*})",
                                  .con = dbcon))

exp_fetch <- dbFetch(exp_query)
dbClearResult(exp_query)

kable(exp_fetch) %>%
  kable_styling(latex_options = c('scale_down'))


```

### Fetch sites

```{r fetch-site, eval = eval_chk}

new_site <- sites$sitename

site_query <- dbSendQuery(dbcon,
                          glue_sql("select * from sites where sitename in ({new_site*})",
                                   .con = dbcon))

site_fetch <- dbFetch(site_query)
dbClearResult(site_query)

kable(site_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

### Fetch treatments

```{r fetch-treat, eval = eval_chk}

new_treat <- treatments$name

treat_query <- dbSendQuery(dbcon,
                           glue_sql("select * from treatments where name in ({new_treat*})",
                                    .con = dbcon))

treat_fetch <- dbFetch(treat_query)
dbClearResult(treat_fetch)

kable(treat_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

### Fetch cultivars

```{r fetch-cultivar, eval = eval_chk}

#TBD

```

### Fetch citations

```{r fetch-citation, eval = eval_chk}

citation_query <- dbSendQuery(dbcon,
                              glue_sql("select * from citations where author = {citations$author} and
                                       year = {citations$year} and
                                       title = {citations$title}",
                                       .con = dbcon))

citation_fetch <- dbFetch(citation_query)
dbClearResult(citation_fetch)

kable(citation_fetch) %>%
  kable_styling(latex_options = c('scale_down'))

```

Close database connection

```{r close-dbcon}
dbDisconnect(dbcon)
```
