# Accessing trait data in R 

## Learning Objectives

In this chapter you will learn:

- How to create a summary of available data to query from a TERRA REF season
- How to query a specific trait
- How to visualize query results


## Introduction

In this chapter, we go over how to query TERRA REF trait data using the [`traits`](https://github.com/ropensci/traits) package. The `traits` package is a way to query for various sources of species trait data, including [BETYdb](https://www.betydb.org/), [NCBI](http://www.ncbi.nlm.nih.gov/), [Coral Traits Disease](http://coraltraits.org/) and others. In this chapter we use BETYdb as our trait source, as it contains the TERRA REF data that we are interested in.

Our example will show how to query for season 6 data <!-- what is a season? need to define this for users> --> and visualize canopy height. In addition to the `traits` package we will also be using some of the [`tidyverse`](https://www.tidyverse.org/) packages, which allow us to maniupulate the data in an efficient, understandable way. If you are unfamiliar with `tidyverse` syntax, we recommend checking out some of the resources [here](https://www.tidyverse.org/learn/).



## Query for available traits

### Getting Started

First, we will need to install and load the traits package from github, and load it into our environment, along with the other packages we will use in this tutorial.

```{r traits-vig-pack, message = FALSE, results = FALSE}
# devtools::install_github('terraref/traits', force = TRUE) # run once
library(traits)
library(ggplot2)
library(lubridate)
library(dplyr)
library(knitr)
```



### Setting options

The function that is used to query BETYdb is called `betydb_query`. To reduce the number of arguments needed to pass into this function, we can set some global options using `options`. In this case, we will set the URL used in the query, and the API version.

```{r traits-vig-bety-opt}
options(betydb_url = "https://terraref.ncsa.illinois.edu/bety/",
        betydb_api_version = 'v1')
```


### Querying available traits

The TERRA REF database contains trait data for many other seasons of observation, and available data may vary by season. Here, we get a visual summary of available traits and methods of measurement for a season.

First we construct a general query for the Season 4 data. This returns all season 4 data. The function `betydb_query` takes as arguments `key = "value"` pairs which represent columns in the database to query. In this example, we `sitename` column for season 4 data, and set the limit to "none" to return all records. By default, the function will search all tables in the database. To specify a particular table you can use the `table` argument.

```{r traits-vig-s4-summary-get-trait, message = FALSE}
# get all of season 4 data 
season_4 <- betydb_query(sitename  = "~Season 4",
                         limit     =  "none")
```

The return value for the `betydb_query` function is just a `data.frame` so we can work with it like any other `data.frame` in R.

Let's plot a time series of all traits returned. First you might notice that the relevant date columns in the `season_4` data.frame are returned as characters instead of a date format. Before plotting, let's get our `raw_date` column into a proper date format using functions from `dplyr` and `lubridate`.


```{r traits-vig-s4-summary-reformat-date, message = FALSE}
season_4 <- season_4 %>%
  mutate(raw_date = ymd_hms(raw_date))
```

### Plot season 4 summary

Now we can create a plot of all of the trait data collected during season 4, including information about the methods used.

```{r traits-vig-s4-summary-plot-data, fig.height = 15, fig.width = 7}
ggplot(data = season_4) +
  geom_point(aes(x = raw_date, y = mean, color = method_name)) +
  geom_line(aes(x = raw_date, y = mean, group = cultivar, color = method_name)) +
  facet_wrap(~trait, ncol = 1, scales = "free_y") +
  xlab("Date") + 
  ylab("Mean trait value") + 
  ggtitle("Season 6 data summary") +
  guides(color = guide_legend(title="Method", ncol = 1, title.position = "top")) +
  theme_bw() +
  theme(legend.position = "bottom")

```


We can view more information about these trait measurements by examining unique values in the trait and trait description columns.

```{r traits-vig-s4-summary-get-trait-info}
traits <- season_4 %>%
  distinct(trait, trait_description)

kable(traits)
```

## Querying a specific trait

### Querying season 6 canopy height data

You may find after constructing a general query as above that you want to only query a specific trait. Here, we query for the canopy height trait by adding the key-value pair `trait = "canopy_height"` to our query function. Note that the limit is also set to return only 250 records, shown here for demostration purposes.

```{r traits-vig-canopy-query, message = FALSE}

canopy_height <- betydb_query(trait     = "canopy_height", 
                              sitename  = "~Season 6",
                              limit     =  250)


```

### Plotting query results

As before, we need to reformat the raw date column. 

```{r traits-vig-fix-date}
canopy_height <- canopy_height %>% 
  mutate(raw_date = ymd_hms(raw_date))
```

And we can generate a time series plot of just the canopy height data.

```{r traits-vig-canopy-plot, warning = FALSE, message = FALSE}

#plot a time series of canopy height 
ggplot(data = canopy_height,
       aes(x = raw_date, y = mean)) +
  geom_point(size = 0.5, position = position_jitter(width = 0.1)) +
  xlab("Date") +
  ylab("Plant height (cm)") +
  ggtitle("Sorghum canopy height, Season 6 TERRA REF") +
  theme_bw()

```



