# Downloading data with BrAPI R package

## Objective: To be able to demonstrate how to get all TERRA REF data available via BrAPI

- Define BrAPI
- Describe `brapi` R package for using API

```{r, eval=FALSE}
library(brapi)
```

# Exploring available data

The function `ba_calls` from `brapi` is supposed to show which functionalities are available for our data base. 

```{r, eval=FALSE}
terraref <- ba_db()$terraref
ba_calls(terraref)
```
This returns no data and some warnings, as below: 

```{shell}
Unknown or uninitialised column: 'datatypes'.Unknown or uninitialised column: 'methods'.Unknown or uninitialised column: 'versions'.No encoding supplied: defaulting to UTF-8.
```

These errors come from two sources: 

1. The GET call using authentication when our server doesn't require any
2. For showing the type of data, our server uses the current API version of the key (`dataTypes`) while the R package uses the older version (`datatypes`)

This can be fixed in the R package code base by doing the following: 

**Remove authorization argument**

This is in `brapiGET` function in the `httr::GET` line. 

*Current*
```{r, eval=FALSE}
  res <- httr::GET(url = url, httr::timeout(25),
        httr::add_headers("Authorization" =
                            paste("Bearer", con$token)))
```

*New*
```{r, eval=FALSE}
  res <- httr::GET(url = url, httr::timeout(25))
```

**Fix case**

This is in `ba_calls` function in the `out$datatypes` line. 

*Current*
```{r, eval=FALSE}
      out$datatypes <- vapply(X = out$datatypes, FUN = paste, FUN.VALUE = "",
                              collapse = "; ")
```

*New*
```{r, eval=FALSE}
      out$datatypes <- vapply(X = out$dataTypes, FUN = paste, FUN.VALUE = "",
                              collapse = "; ")
```

Note: the latter does break for the example database used for BrAPI, sweetpotatobase, because the server still uses `datatypes`.

Once these two changes are made, they can implemented locally by rebuilding the `brapi` R package by running `devtools::install(".")` from inside the `brapi` folder. 
