---
title: "Accessing Sensor Metadata"
subtitle: "Response Curves and other Metadata"
author: "David LeBauer, Craig Willis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(jsonlite)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())


```

# Introduction

This tutorial will demonstrate how to access sensor metadata from within R. All of the sensor metadata is public, and can be queried via the API using the url `https://terraref.ncsa.illinois.edu/clowder/api/datasets/<id>/metadata.jsonld`.

For further information about sensor metadata see the [Sensor Data Standards](/sensor-data-standards.md) section.

## Avaialble Sensor metadata


### Example: RSR curves for PAR, PSII and NDVI

* par: https://terraref.ncsa.illinois.edu/clowder/api/datasets/5873a8ce4f0cad7d8131ad86/metadata.jsonld
* pri: https://terraref.ncsa.illinois.edu/clowder/api/datasets/5873a9174f0cad7d8131b09a/metadata.jsonld
* ndvi: https://terraref.ncsa.illinois.edu/clowder/api/datasets/5873a8f64f0cad7d8131af54/metadata.jsonld

```{r}
sensor_datasets <- jsonlite::fromJSON("data/sensor_metadata_datasets.json")
knitr::kable(sensor_datasets[,c(2,3,1)])
```

### Looking at Skye NDVI meta-data

```{r}

ndvi_metadata <- jsonlite::fromJSON("https://terraref.ncsa.illinois.edu/clowder/api/datasets/581787524f0ce77b6655b2c7/metadata.jsonld")

knitr::kable(t(ndvi_metadata$content[-21]), col.names = '')

```
#### RSR curve 

```{r rsr-curve}
ndvi_up_rsr <- cbind(direction = 'up',
                     bind_rows(lapply(ndvi_metadata$content$rsr$up, unlist)))
ndvi_down_rsr <- cbind(direction = 'down', 
                    bind_rows(lapply(ndvi_metadata$content$rsr$down, unlist)))
ndvi_rsr <- rbind(ndvi_up_rsr, ndvi_down_rsr) 
ggplot(data = ndvi_rsr, aes(x = wavelength, linetype = direction), alpha = 0.4) +
  geom_line(aes(y = ch1), color = 'green') + 
  geom_line(aes(y = ch4), color = 'red') +
  ylab('relative response')
```

