---
title: "Solar Radiance"
author: "David LeBauer"
date: "March 9, 2018"
output: html_document
---


# VNIR Radiometer Data

An Ocean Optics STS Spectrometer measures downwelling solar spectral radiance every 5s on top of the Gantry.

Lets look at the output from this sensor over the course of the day (and then we will see how to acces the data):

```{r one-downwelling-spectra}
library(tidyverse)
library(ggridges)
load('data/spectra.RData')

s <- spectra_long %>% 
  mutate(hour = hour, 
         radiance = radiance - min(radiance)) %>% 
  arrange(hour, wavelength) 
  # subset for faster exploration: %>% slice(1:(24*64)*16) 

ggplot(data = s, aes(x = wavelength, y = hour, group = hour, 
                                height = radiance )) +
  geom_density_ridges(stat = 'identity', scale = 6, size = 0.25, alpha = 0.7, color = 'white') +
  theme_ridges(grid = FALSE, center_axis_labels = TRUE) +
  scale_y_continuous(trans = 'reverse') +
  ggtitle("Downwelling Spectral Radiance", 
          subtitle = "hourly spectra from April 15, 2017")

# Fun Challenge: implement wavelength --> color mapping
#scale_color_gradientn(
#    colors = c('white', 'purple', 'blue', 'cyan', 'green', 'yellow', 'orange', 'red', #'black'), 
#    values = c(300, 420, 570, 530, 580, 620, 700, 800))

```

## Query from Environmental logger netCDF files

There are 20 observations of 1024 individual wavelengths_per minute_ = `r 20 * 60 * 24 * 1024` data points per day. We convert these data to CF standards and store them in netCDF file formats.

These are used in the hyperspectral workflow. 

Let's take a look at one of these files:

```{r netcdf-metadata}
library(tidyverse)
library(ncdf4)
library(udunits2)
library(lubridate)
if(!require(tidync)){
  devtools::install_github('hypertidy/ncmeta')
  devtools::install_github('hypertidy/tidync')
} 

library(ncmeta)
library(tidync)

envlog_file <- "/data/terraref/sites/ua-mac/Level_1/envlog_netcdf/2017-08-21/envlog_netcdf_L1_ua-mac_2017-08-21.nc"
envlog.nc <- nc_open(envlog_file, readunlim = TRUE)

time <- envlog.nc$dim$time$vals
wvl <- envlog.nc$dim$wvl_lgr$vals

metadata <- envlog.nc$var %>% bind_cols()

#flx_dwn <- ncvar_get(envlog.nc, ')

#if(!require(ncdf4.helpers)) install.packages("ncdf4.helpers")
#ts <- ncdf4.helpers::nc.get.time.series(envlog.nc)
s <- tidync::tidync(envlog_file)
  
  
nc_metadata <- ncmeta::nc_meta(envlog_file)

nc_metadata$variable %>% 
  select(name, longname, units, ndims) %>% 
  filter(!grepl('raw', name)) %>% 
  knitr::kable()
```



Now lets query the Downwelling Spectral Irradiance  (flx_spc_dwn) from this file:

```{r}

library(tidync)
flx_spc_dwn <- ncvar_get(envlog.nc, 'flx_spc_dwn', )
time <- ncvar_get(envlog.nc, 'time')
dim(flx_spc_dwn)

if(!require(rasterVis)) install.packages("rasterVis")

library(rasterVis)
gplot(flx_spc_dwn) + 
  geom_tile(aes(fill = value))
  
```


```{r}
time = flx_spc_dwn$  radiance = as.vector(flx_spc_dwn)
get_spectra <- function(date, site = 'ua-mac'){
  
  
  envlog_file <- file.path("/data/terraref/sites", site, "Level_1/envlog_netcdf", date, 
                           paste0("envlog_netcdf_L1_", site, "_", date, ".nc"))
  envlog.nc <- nc_open(envlog_file, readunlim = FALSE)
  timepoints <- ud.convert(4*1:5, 'h', '5s')
  
  flx_spc_dwn <- ncvar_get(envlog.nc, 'flx_spc_dwn')[,timepoints]
  datetime <- ymd("1970-01-01", tz ="America/Phoenix") + 
    seconds(ud.convert(ncvar_get(envlog.nc, 'time'), 'day', 's'))
  datetime[timepoints]
  
  z <- as.data.frame(flx_spc_dwn)
  colnames(z) <- datetime[timepoints]
  
  wvl <- ncvar_get(envlog.nc, 'wvl_lgr')
  wvl.idx <- sapply(c(34:81*10), 
                    function(x) which.min(abs(x-wvl)))
  zz <- z[wvl.idx,]
  rownames(zz) <- round(wvl[wvl.idx])
  return(t(zz))
  write(zz, file = paste0('tmp/',date, '.csv'))
}
  

dates <- seq(ymd('2017-04-15'), ymd('2017-11-12'), by = '10 days')
all_spectra <- lapply(as.character(dates), get_spectra)
spectra.df <- lapply(all_spectra, as.data.frame)
zzz <- do.call(rbind, unname(spectra.df))
write.csv(zzz, '~/tmp/spectra.csv')
return(list(spc = spc, wvl = wvl, date = ymd(strftime(datetime, '%Y%m%d')), datetime = datetime))



for(date in c('2016-06-21', '2016-09-21', '2016-12-21', '2017-03-21', '2017-05-21')){
  
  directory <- file.path("/data/terraref/sites/ua-mac/Level_1/envlog_netcdf/", date)
  files <- dir(directory, full.names = TRUE)  
  spectra_list <- lapply(files, function(x){
    metnc <- nc_open(x)
    spc <- ncvar_get(metnc, 'flx_spc_dwn')
    datetime <- ymd("1970-01-01") + 
      seconds(ud.convert(ncvar_get(metnc, 'time'), 'day', 's'))
    wvl <- ncvar_get(metnc, 'wvl_lgr')
    time <- hour(datetime) + 
      minute(datetime)/60 + 
      second(datetime)/3600
    return(list(spc = spc, wvl = wvl, date = ymd(strftime(datetime, '%Y%m%d')), datetime = datetime))
    
  })
  
  spectra_df <- do.call('cbind',(lapply(spectra_list, '[[', 'spc') ))
  dim(spectra_df)
  
  time <- do.call('c',lapply(spectra_list,'[[','datetime'))
  wavelengths <- spectra_list[[1]]$wvl
  save(spectra_df, time, wavelengths, file = file.path('data', paste0("spectra",date,".Rdata")))
  idx <- 1+0:700*24
  i <- 1:length(hr)[!is.na(hr)]
  library(lubridate)
  hr <- hour(time) + minute(time)/60 + second(time)/3600
  png(filename = paste0('data/spectra',date,'.png'))
  image(x = wavelengths, y = as.numeric(hr[idx]), spectra_df[,idx],
        ylab = 'hour of day', 
        xlab = 'wavelength (nm)',
        col = cm.colors(n=100),zlim = c(-1,2.1),
        main = paste0('diurnal solar spectral radiation\n',date))
  dev.off()
  
}  
library(lubridate)
library(data.table)
library(udunits2)

time <- ncvar_get(metnc, 'time')

wavelengths <- ncvar_get(metnc, 'wvl_lgr')

f_down_spectrum <- ncvar_get(metnc, 'flx_spc_dwn')

library(ggplot2)

ggplot() + 
  geom_point(aes(wavelengths, f_down_spectrum[,1])) +
  geom_line(aes(wavelengths, f_down_spectrum[,1]))

f_down_means <- rowMeans(f_down_spectrum)

ggplot() + 
  geom_point(aes(wavelengths, f_down_means)) +
  geom_line(aes(wavelengths, f_down_means))

print(metnc)

```

### Your turn:

Can you see the effect of the August 21, 2017 solar eclipse on the diurnal spectral radiance?

## Raw sensor data

Here we can found the original data written by the sensor. Unlike above, these are in text files and are not in a standard format like the CF format above.

```{r raw-met, cache=TRUE}
metfile <- "/data/terraref/sites/ua-mac/raw_data/EnvironmentLogger/2017-05-31/2017-05-31_12-19-38_environmentlogger.json"
met <- jsonlite::fromJSON(metfile)
writeLines(jsonlite::toJSON(met), con = file('foo.json'))

timestamp <- lubridate::ymd_hms(met$environment_sensor_readings$timestamp)

wavelengths <- met$environment_sensor_readings$spectrometer$wavelength[[1]]

spectra <- do.call('rbind', met$environment_sensor_readings$spectrometer$spectrum)

library(dplyr)
spectra <- do.call('rbind', met$environment_sensor_readings$spectrometer$spectrum)

#colnames(spectra) <- wavelengths
#rownames(spectra) <- met$environment_sensor_readings$timestamp
image(x = timestamp, y = wavelengths, z = spectra)
```


```{r}
library(dplyr)
library(readr)
date = '2017-04-15'
load_loggerdata <- function(date){
  path <- file.path("/data/terraref/sites/ua-mac/raw_data/EnvironmentLogger", date)
  files <- dir(path, full.names = TRUE)
  loggerdata <- lapply(files, jsonlite::fromJSON) 
  timestamp <- combine(sapply(loggerdata, function(x){
    t <- x$environment_sensor_readings$timestamp
    lubridate::ymd_hms(t)
  }))
  return(list(data = loggerdata, timestamp = timestamp))
}

extract_downwelling_irradiance <- function(logdata){

  wavelengths <- logdata$data[[1]]$environment_sensor_readings$spectrometer$wavelength[[1]]
  
  spectra <- do.call('rbind', lapply(logdata$data, function(x){
    do.call('rbind', x$environment_sensor_readings$spectrometer$spectrum)
    }
  ))
  # image(x = timestamp, y = wavelengths, z = spectra)
  return(list(spectra = spectra, wavelengths = wavelengths, timestamp = logdata$timestamp))
}

extract_logger_met <- function(logdata){
  
  met <- do.call('rbind', lapply(logdata$data, function(x){
    tmp_met <- x$environment_sensor_readings
    data.frame(par = tmp_met$`sensor par`$value,
               co2 = tmp_met$`sensor co2`$value,
               sundir = tmp_met$weather_station$sunDirection$value,
               pressure = tmp_met$weather_station$airPressure$value,
               brightness = tmp_met$weather_station$brightness$value,
               rh = tmp_met$weather_station$relHumidity$value,
               temp = tmp_met$weather_station$temperature$value,
               wind_dir = tmp_met$weather_station$windDirection$value,
               wind_speed = tmp_met$weather_station$windVelocity$value)
    
  })) 
  return(met)
}

env_log_data <- load_loggerdata(date = '2017-04-15')
env_log_spectra <- extract_downwelling_irradiance(env_log_data)
env_log_met <- extract_logger_met(env_log_data)

```

#### Plots

```{r}
library(lubridate)
library(dplyr)
library(tidyr)
time <- env_log_data$timestamp

hourly_index <- 1+0:23*720

time_hr <- time[hourly_index]
hourly_spectra <- env_log_spectra$spectra[hourly_index,]
wavelengths <- env_log_spectra$wavelengths

colnames(hourly_spectra) <- wavelengths

image(x = time_hr, y = wavelengths, z = hourly_spectra, 
      xlab = 'local time', ylab = 'wavelength (nm)')
```

```{r spectra-ggplot}

spectra_df <- data.frame(hour = 1:24, hourly_spectra)

spectra_long <- spectra_df %>% 
  gather(key = wavelength, value = radiance, -hour) %>% 
  mutate(wavelength = as.numeric(gsub("X", "", wavelength))) 

colnames(spectra_long)

library(ggplot2)
ggplot(data = spectra_long, aes(x = wavelength, y = radiance)) +
  geom_line(size = 0.1) +
  ggthemes::theme_tufte() +
  facet_wrap(~hour, ncol = 6) +
  ggtitle(paste('spectra on', date))

```
