---
title: "Getting Started with BETYdb"
author: "David LeBauer"
date: "`r Sys.Date()`"
output: html_document
---

## TERRA Ref Trait Database

The TERRA Ref program uses the BETYdb database and web application software to store plant and plot level trait data. 

### BETYdb: database software and web application

The BETYdb software is actively used and developed by the [TERRA Reference](terraref.org) program as well as by the [PEcAn project](pecanproject.org).

For more information about BETYdb, see the following:

* BETYdb documentation (available via the web application under 'Docs')
  * _Data Access_
  * _Data Entry Workflow:_ how to add data to the database
  * _BETYdb Technical Documentation_ is written for advanced users and website and database administrators who may also be interested in the [full database schema](betydb.org/schemas)
* BETYdb: A Yield, Trait and Ecosystem Service Database Applied to Second Generation Bioenergy Feedstocks. ([LeBauer et al, 2017](dx.doi.org/10.1111/gcbb.12420))

The TERRA REF trait database (terraref.ncsa.illinois.edu/bety) uses the BETYdb data schema (structure) and web application.
There are at least a half-dozen other databases using the BETYdb software that these exercises will work with, though the results will depend on the available data.
The first, betydb.org is described in LeBauer et al, 2017.
Others are listed in the 'distributed BETYdb' section of the technical documentation.

One database, terraref.ncsa.illinois.edu/terra-test, houses a simulated dataset that is used in [lesson 1: A simulated data set](../traits/01-simulated-sorghum.Rmd) and does not require an account to access the data.
BETYdb is only designed to keep the primary data private. Metadata such as field management and experimental design are available if the url is public.

## Getting an account for the TERRA trait database

* sign up for an account at terraref.ncsa.illinois.edu/bety
* sign up for alpha user [link to form]
* wait for database access to be granted
* Your API key will be sent in the email; it can also be found - and regenerated - by navigating to 'data --> users' in the web interface

TODO add signup info from handout

## First steps: download data from web interface

TODO add steps to download csv from the web interface

Note that the web interface only provides a core set of data. More complex queries, such as those in the [Agronomic metadata](../traits/04-agronomic-metadata.Rmd)

## Advanced: Using URLs to construct Queries

The first step toward reproducible pipelines is to automate the process of searching the database and returning results. This is one of the key roles of an Application programming interface, or 'API'. You can learn to use the API in less than 20 minutes, starting now. 

### What is an API?

An API is ...

### Using Your API key to Connect

An API key is like a password. It allows you to access data, and should be kept private. 
Therefore, we are not going to put it in code that we share. The one exception is the key 9999999999999999999999999999999999999999 that will allow you to access metadata tables (all tables except _traits_ and _yields_). 

A common way of handling private API keys is to place it in a text file in your home directory. 
Don't put it in a project directory where it might be inadvertently shared.

Here is how to find and save your API key:

* click file --> new --> text file
* copy the api key that was sent when you registered into the file
* file --> save as '~/.betykey'

Equivalently in R `r writeLines('9999999999999999999999999999999999999999', con = '~/.betykey')` or at the command line `sh echo 9999999999999999999999999999999999999999 > ~/.betykey`

For the purposes of the tutorial, you can assign it to the `betykey` variable in the console window.

### Constructing a URL query

First, lets construct a query by putting together a URL.

1. start with the database url: `terraref.ncsa.illinois.edu/bety`
  * this url brings you to the home page
2. Add the path to the API, `/api/beta`
  * now we have terraref.ncsa.illinois.edu/bety/api/beta, which points to the API documentation
3. Add the name of the table you want to query. Lets start with `variables`
  * terraref.ncsa.illinois.edu/bety/api/beta/variables
4. add query terms by appending a `?` and combining with `&`, for example:
  * `key=9999999999999999999999999999999999999999`
  * `type=trait` where the variable type is 'trait'
  * `name=~height` where the variable name contains 'height'
5. This is your complete query:
  * `terraref.ncsa.illinois.edu/bety/api/beta/variables?type=trait&name=~height&key=9999999999999999999999999999999999999999`
  
**Your Turn**

> write a URL that will query the database for sites with "Field Scanner" in the name field. Hint: combine two terms with a `+` as in `Field+Scanner`

What do you see? Do you think that this is all of the records? What happens if you add `&limit=none`? 

## Using the R traits package to query the database

The rOpenSci traits package makes it easier to query the TERRA REF trait database, or any database that uses BETYdb software.

First, make sure we have the latest version 

```{r install_traits, echo=FALSE}
if(packageVersion("traits") == '0.2.0'){
  devtools::install_github('ropensci/traits')
}
```


```{r setup}
library(traits)
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(ggplot2)
library(ggthemes)
theme_set(theme_bw())
library(dplyr)
```



```{r writing-key}
# This should be done once with the key sent to you in your email
# writeLines('abcdefg_rest_of_key_sent_in_email', 
#            con = '~/.betykey')

# Example with the public key:
writeLines('9999999999999999999999999999999999999999', 
           con = '~/.betykey_public')
```
```{r traits}
terraref_test_url <- "https://terraref.ncsa.illinois.edu/bety-test/"

betykey <- readLines('~/.betykey_public', warn = FALSE)

```

```{r simulated-LAI}
sorghum_lai <- betydb_query(table = 'search',
                            trait = "LAI",
                            api_version = 'beta',
                            limit = 5000,
                            betyurl = terraref_test_url, 
                            key = betykey)

#traits <- betydb_query(table = "traits",
#              api_version = 'beta',
#              limit = 5000,
#              betyurl = terraref_test_url, 
#              key = betykey)



```

```{r}

ggplot(data = sorghum_lai) +
  geom_smooth(aes(x = lubridate::yday(lubridate::ymd_hms(raw_date)), y = mean, color = as.factor(lubridate::year(lubridate::ymd_hms(raw_date)))), span = 0.5) +
#  scale_x_datetime(date_breaks = '6 months', date_labels = "%b %Y") +
  ylim(c(0,6)) + 
  ylab("Day of Year") + xlab("Leaf Area Index") +  
  labs(color='Year')

```

