---
title: "Can high throughput phenotyping produce data like this?"
author: "David LeBauer"
date: "February 1, 2015"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(ggplot2)
library(ggthemes)
library(GGally)
theme_set(theme_bw())

```

# Background: The design of a simulated dataset

To explore the potential of phenotyping data, we have simulated the type of data that might be observed by daily scans by drone or robot. 

We have simulated 500 genotypes across 12 sites and five years using a mechanistic model. 

_Note that these data sets contain numerical artifacts and scientific reinterpretations for illustrative purpose._

All of these simulated datasets are released with an unrestrive [copyright](https://creativecommons.org/publicdomain/zero/1.0/).  This means you can copy, modify, ans share the data.  Please keep in mind that the data sets are not production quality - they have been developed solely to inspire and solicit feedback.

### A note on variable names

I have used the variable names currently used in BETYdb.org/variables, along with names inspired by the more standardized naming Climate Forecasting conventions. However, this is a very early pre-release, and we welcome comments on how such data should be formatted and accessed can be discussed on GitHub.

> Exercise: can you locate the relevant issues?

[This is a slideshow](https://docs.google.com/presentation/d/10aN_5whs8y9SOC8Y9Rj1kbWCfi7YG3yFyNqVX4JUr2U/edit?usp=sharing) of interfaces from the broader community that could serve as a common interface.

# Design of Simulation Experiment

500 Sorghum lines grown at each of three sites, four blocks per site, along a N-S transect in Illinois over five years (2021-2025). 

## Time Span (2021-2025) 

These are historic data, but the years have been changed to emphasize the point that these are not real data. The years have been chosen to select climate extremes. Two years were dry, two were wet, and one was average.

| year | drought index |
|-----|-----|
| 2021 | wet |
| 2022 | dry |
| 2023 | normal |
| 2024 | wet |
| 2025 | dry |


### Sites

These are approximate locations used to query the meteorological and soil data used in the simulations. 
 
| site name | latitude | longitude | 
|-------|------|------|
| north | 42.0 | -88.5 |
| central | 40.0 | -88.5 |
| south | 37.0 | -88.5 |

Each site has four replicate fields: A, B, C, D. This simulated dataset assumes each field within a site has similar, but different meteorology (e.g., as if they were all in the same county). 

```{r query-sites}
sites <- betydb_query(table = 'sites', name = 'Season+2')
# plot polygons, how? leaflet? something simple?
```

> Your Turn: How would you add these points to a map?

## Meteorology

| variable_id|name                           |standard_name       |units                          |description                                                                                                                                                                                                      |
|-----------:|:------------------------------|:---------------|:------------------------------|:----------------|
|            | Tmin                               | | C  |Daily max temperature |
|            | Tmax                               |  | C  |Daily min temperature |
|            | Tavg                               |  | C  |Daily mean temperature |
|            | precipitation                    | precipitation_flux | mm/d                        |Daily precipitation |


## Genotypes

Two-hundred and twenty-seven lines were grown at each site. Each line is identified by a unique integer in the range [9915:10141]

## Phenotypes

The phenotypes associated with each genotype is in the file `phenotypes.csv`. 

These 'phenotypes' are used as input parameters to the simulation model. We often refer to these as 'traits' (as opposed to biomass or growth rates, which are states and proceses). In this example, we assume that 'phenotypes' are time-invariant.


| variable_id|name                           |standard_name                                                 |units                          |Description                                                                                                                                                                                                      |
|-----------:|:------------------------------|:-------------------------------------------------------------|:------------------------------|:----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
|          |genotype                 |                            |               |genetically and phenotypically distinct line |         
|          |Vmax                 |                            | umol m-2 s-1              | maximum carboxylation of Rubisco according to the Collatz model  |      
|          38|cuticular_cond                 |conductance_of_fully_closed_stomata                           |umol H2O m-2 s-1               |leaf conductance when stomata fully closed                                                                                                                                                                       |
|          15|SLA                            |specific_leaf_area                                            |m2 kg-1    |Specific Leaf Area|
|          39|quantum_efficiency             |mole_ratio_of_carbon_dioxide_to_irradiance_in_leaf  |fraction   |see Farqhuar model |
|          18|LAI                            |leaf_area_index          |m2 leaf m-2 ground |Leaf Area Index |
|          31|c2n_leaf |mass_ratio_of_carbon_to_nitrogen_in_leaf   |ratio  |C:N ratio in leaves|
|         493|growth_respiration_coefficient |respiration_coefficient_for_growth  |mol CO2 / mol net assimilation |amount of CO2 released due to growth per unit net photosynthesis  |
|           7|leaf_respiration_rate_m2       |respiration_rate_per_unit_area_in_leaf                        |umol CO2 m-2 s-1               |Not really ""dark respiration"" Often this is respiration that occurs in the light. Date and time fields ""should"" identify pre-dawn (nightime/dark) leaf resp vs the Rd that comes from a A-Ci or A-PPFD curve |
|           4|Vcmax                          |rubisco_carboxylation_rate_in_leaf_assuming_saturated_rubp    |umol CO2 m-2 s-1               |maximum rubisco carboxylation capacity  |
|         404|stomatal_slope.BB              |stomatal_slope_parameter_assuming_ball_berry_model            |ratio      |slope parameter for Ball-Berry Model of stomatal conductance |
|           5|Jmax   |electron_transport_flux_in_thylakoid_assuming_saturated_light |umol photons m-2 s-1   |maximum rate of electron transport  |
|         492|extinction_coefficient_diffuse |extinction_coefficient_for_diffuse_light_in_canopy            |                               |canopy extinction coefficient for diffuse light                                                                                                                                                                  |

## Simulated Sensor Data

This dataset includes what a sensor might observe, daily for five years during the growing season.

| variable_id|name                           |standard_name       |units                          |Description                                                                                                                                                                                                      |
|-----------:|:------------------------------|:---------------|:------------------------------|:----------------|
|            | sitename                     |                                                              |                        |  Name of site |
|            | plotid                               |                                                              |                        | experimental replicate plot |
|            |year                             |                                                              |                        | |
|            | date                           | | YYYY-MM-DD         | |
|            |  Stem                             |  stem_biomass_content  |                        | Mg / ha |
|            |  Leaf                             |  leaf_biomass_content  |                        | Mg / ha |
|            |  Root                             |  root_biomass_content  |                        | Mg / ha |
|            |  Rhizome                             |  rhizome_biomass_content  |                        | Mg / ha |
|          18|LAI                            |leaf_area_index   |ratio             |  Leaf Area Index is the ratio of leaf area to ground area |
|          |NDVI                            |normalized_difference_vegetation_index                                               |ratio             | commonly used vegetation index |
|            | Height                               |   canopy_height | m                        | |


# Quick start


For simplicity, and because I neither have a model nor the data to simulate _Sorghum_, I have started with some phenotypes and simulations of plant growth based on a model that simulates the growth of biomass crops including Miscanthus, Switchgrass, Sugarcane, and coppice Willow.

## Some background / methods.

I start with simulation of Miscanthus, over Illinois as a proxy for Sorghum, since I have a model, [BioCro, Miguez et al, 2009](github.com/ebimodeling/biocro), that simulates Miscanthus. [The code used to run the model and add noise is on GitHub](https://github.com/ebimodeling/biocro_regional/edit/master/vignettes/regional_pecan_workflow.Rmd). 
Like Sorghum, Miscanthus uses C4 photosynthesis, and this is used to compute carbon uptake at hourly time steps in the simulation model. 
Unlike Sorghum, Miscanthus grows clonally and is propagated by Rhizome instead of by seed. Furthermore, Miscanthus is perennial: it re-grows each year from carbon stored in rhizomes.
The 'genotypes' are based on five-hundred quasi-random parameterizations of a biophysical crop model.

The virtual "Miscanthus" is encoded as a set of associated species and prior estimates of the phenotypes used to parameterize the simulation model. The concept of Plant Functional Type (PFT) is any group of one or more plants, originally functionally related species such as 'C4 crops' or 'hardwood trees'. We use PFT at the finest scale of genotype. In the actual research that generated these data, this was the clone _Miscanthus x giganteus_, and this is [PFT #123 in BETYdb](https://www.betydb.org/pfts/123). 

## A Simulated Sorghum Breeding Population

In make-believe land we take a bunch of data generated above and start to rename things

* run_id -> genotype #represents quasi-random set of traits
* lat, lon -> To estimate "E"
   * for within site E effects, use a few points within 1/4 degree lat/lon 
   * for across site E effects, use southern, central, and northern Illinois.

### Accessing the TERRA Simulated Data Database


```{r}

library(traits)

terraref_test_url <- "https://terraref.ncsa.illinois.edu/bety-test/"
## note that this key accesses public data. In part 2 you will use your own key to access the actual data

public_key <- "9999999999999999999999999999999999999999"

traits <- c("Vcmax", "c2n_leaf", "cuticular_cond", "SLA", "quantum_efficiency", "leaf_respiration_rate_m2", "stomatal_slope.BB", "Jmax", "chi_leaf", "extinction_coefficient_diffuse")


SLA <- (betydb_search(query = "SLA", betyurl = terraref_test_url, key = public_key)
  %>% filter(scientificname == 'Sorghum bicolor'))

```


Now we have queried all of the Specific Leaf Area records, we can see that these data are summarized at the level of genotype. 


```{r sla-fig, fig.height=2, fig.width=6}
ggplot(SLA) + 
  geom_histogram(aes(mean), binwidth = 1)
```

## Your turn: query the list of available traits from the variables table

```{r query-traits}


traits <- betydb_search("SLA Vcmax",  betyurl = terraref_test_url, key = public_key)

p <- phenotypes %>% dplyr::select(Vcmax, c2n_leaf, cuticular_cond, SLA, quantum_efficiency,
                                  leaf_respiration_rate_m2, stomatal_slope.BB, growth_respiration_coefficient,
                                  Jmax, chi_leaf, extinction_coefficient_diffuse) %>% setDT
                                
ggpairs(p,  
        lower = list(continuous = 'density'),
        upper = list(continuous = 'blank'),
        axisLabels = 'internal')
        
```

### Simulated Biomass

```{r}

# query some time series of yield, LAI, 
# time series and scater plots of 

```


```{r}
library(data.table)
library(dplyr)

phenotypes <- fread("phenotypes.csv")
sorghum_simulation <- fread("sorghum_simulation.csv")
library(ggplot2)
library(ggthemes)
library(GGally)
theme_set(theme_tufte())

p <- phenotypes %>% dplyr::select(Vcmax, c2n_leaf, cuticular_cond, SLA, quantum_efficiency,
                                  leaf_respiration_rate_m2, stomatal_slope.BB, growth_respiration_coefficient,
                                  Jmax, chi_leaf, extinction_coefficient_diffuse) %>% setDT
                                
ggpairs(p,  
        lower = list(continuous = 'density'),
        upper = list(continuous = 'blank'),
        axisLabels = 'internal')

```

### Time series


```{r}
library(lubridate)
library(data.table)
# need to query from API
s <- sorghum_simulation %>% 
  mutate(Day = yday(date), 
         line = as.factor(paste('Sim', genotype)),
         Biomass = Stem + Leaf) %>% setDT

## figure out best / worst
s[year == 2023][which.max(Stem)]
s[year == 2023 & month(date) == 10][which.min(Stem)]
theme_set(theme_tufte()) 
ggplot() + 
  geom_smooth(data = s[genotype >9950 & genotype < 10000],
              aes(x = Day, y = Biomass, fill = NULL, group = genotype), alpha = 0.1, color = 'grey') +
  geom_smooth(data = s[genotype == 10020],
              aes(x = Day, y = Biomass), color = 'darkgreen') +
  geom_smooth(data = s[genotype == 9934],
              aes(x = Day, y = Biomass), color = 'red') +
  facet_wrap(~year, nrow = 1) 
        
```

# How to obtain data and give feedback:

Please provide feedback by leaving a comment below, commenting on the issue in our [GitHub repository](https://github.com/terraref/reference-data/issues/20), emailing me, [dlebauer@illinois.edu](mailto:dlebauer@illinois.edu), or visiting the [TerraRef Reference Data chatroom](https://gitter.im/terraref/reference-data).

* If you do something cool, please send comments and figures! 
* I also can provide similar data at hourly or higher frequency as well as other processes and environmental drivers, 
Data are located on Box: https://uofi.box.com/sorghum-simulation

## Data on BETYdb

These data have been uploaded to a [test instance of our database, BETYdb.](http://terraref.ncsa.illinois.edu/bety-test/search?search=sorghum) and can be accessed in the following ways:

1. web interface: search + download
 * go to http://terraref.ncsa.illinois.edu/bety-test and search for "Sorghum" in the search box
 * click "Download Results"
2. API: 
  * basic search: http://terraref.ncsa.illinois.edu/bety-test/search?search=sorghum
  * Draft API Documentation: https://terraref.ncsa.illinois.edu/bety-test/api/docs
3. SQL Dump: https://terraref.ncsa.illinois.edu/bety-test/dump/bety.tar.gz
4. Other Methods: https://pecan.gitbooks.io/betydb-data-access/content/introduction.html
