---
title: "Accessing Trait Data in R"
author: "David LeBauer"
output: html_document
---


## Using the R traits package to query the database

The rOpenSci traits package makes it easier to query the TERRA REF trait database, or any database that uses BETYdb software.

First, make sure we have the latest version from the terraref fork of the repository on github. (you can install using the standard `install.packages('traits')` but I can't promise everything will work.

### Install the package

```{r install_traits, echo=FALSE}
devtools::install_github('terraref/traits')
```

Now, we can load the packages that we will need to get started.

```{r 00-setup}
library(traits)
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(ggplot2)
library(ggthemes)
theme_set(theme_bw())
library(dplyr)
```



```{r writing-key}
# This should be done once with the key sent to you in your email
# writeLines('abcdefg_rest_of_key_sent_in_email', 
#            con = '~/.betykey')

# Example with the public key:
writeLines('9999999999999999999999999999999999999999', 
           con = '~/.betykey_public')
```

#### R - using the traits package

The R traits package is an API 'client'. It does two important things:
1. It makes it easier to specify the query parameters without having to construct a URL
2. It returns the results as a data frame, which is easier to use within R

Lets start with the query of information about Sorghum from species table from above

```{r query-species}

sorghum_info <- betydb_query(table = 'species',
                            genus = "Sorghum",
                            api_version = 'beta',
                            limit = 'none',
                            betyurl = "https://terraref.ncsa.illinois.edu/bety/", 
                            key = readLines('~/.betykey', warn = FALSE))

```

#### R - setting options for the traits package

Notice all of the arguments that the `betydb_query` function requires? We can change this by setting the default connection options thus:


```{r}
options(betydb_key = readLines('~/.betykey', warn = FALSE),
        betydb_url = "https://terraref.ncsa.illinois.edu/bety/",
        betydb_api_version = 'beta')
```

Now the same query can be reduced to:

```{r sv_area}
sorghum_height <- betydb_query(table = 'search',
                               trait = "plant_height",
                               site  = "~MAC",
                               limit = 'none')
```

### Time series of height

Now we can take a look at the data that we have just queried. 

```{r}
ggplot(data = sorghum_height, 
       aes(x = lubridate::yday(lubridate::ymd_hms(raw_date)), y = mean, color = cultivar)) +
  geom_smooth(se = FALSE, size = 0.5) +
  geom_point(size = 0.5, position = position_jitter(width = 0.1)) + 
#  scale_x_datetime(date_breaks = '6 months', date_labels = "%b %Y") +
#  ylim(c(0,6)) + 
  xlab("Day of Year") + ylab("Plant Height") + 
  guides(color = guide_legend(title = 'Genotype')) +
  theme_bw()

```

