---
title: "Plot level data from the field scanner in Maricopa, AZ"
author: "David LeBauer, Chris Black"
date: "`r Sys.Date()`"
output: md_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(traits)
library(rgeos)
library(sp) # for implicitly called rbind.SpatialPolygons method
library(leaflet)


betyurl <- "https://terraref.ncsa.illinois.edu/bety/"
betykey <- "9999999999999999999999999999999999999999"
```


## Plots


query sites for season 2

```{r}
sites <- betydb_query(
	table = "sites",
	city = "Maricopa", sitename = "~Season 2 range",
	limit = "none",
	betyurl = betyurl, key = betykey, api_version = "beta")
```

plot polygons on a map

```{r}
site_bounds <- (sites
	%>% rowwise()
	%>% do(boundaries = readWKT(text = .$geometry, id = .$id)))

site_bounds <- do.call('rbind', site_bounds$boundaries)

leaflet() %>% addTiles() %>% addPolygons(data=site_bounds)
```

## Cultivars

```{r}
cultivars <- (
	betydb_query(
		table = "cultivars", limit = "none",
		betyurl = betyurl, key = betykey, api_version = "beta")
	%>% rename(cultivar_id = id))

cultivars_traits <- (cultivars
	%>% group_by(cultivar_id)
	%>% do(trait_id = betydb_record(
			id = .$cultivar_id, table = "cultivars",
			betyurl = betyurl, key = betykey, api_version = "beta"
			)$traits$trait.id)
	%>% unnest())

sites_traits <- (sites
	%>% group_by(id)
	%>% do(trait_id = betydb_record(
			id = .$id, table = "sites",
			betyurl = betyurl, key = betykey, api_version = "beta"
			)$traits$trait.id)
	%>% unnest())

sites <- (sites
	%>% left_join(sites_traits, by = "id")
	%>% left_join(cultivars_traits, by = "trait_id")
	%>% left_join(cultivars, by = "cultivar_id")
	%>% unique())

```


## Time series of canopy cover, height, NDVI

First look up variables by name. Let's look for measurements related to canopy size:

```{r}
variables <- betydb_query(
		table = "variables", name = "~^(NDVI|canopy_height|canopy_cover|)$",
		betyurl = betyurl, key = betykey, api_version = "beta")

variables %>% select(id, name, units, n_records = `number of associated traits`)
```

Exercise: Why are there two variables named canopy_height, and what database fields should you examine to decide which one you want?

Now retrieve all available measurements for each variable.

```{r}
vars_measures <- (variables
	%>% group_by(id, name)
	%>% do(traits = betydb_record( # Get full trait list by variable ID
		id = .$id, table = "variables",
		betyurl = betyurl, key=betykey, api_version = "beta"
		)$traits))

# Only needed if some variables may contain zero traits
# If none are empty, can just do `vars_measures %>% unnest()`
traitdata <- left_join( # ensures we keep a blank row for any variables with no trait info
	vars_measures %>% select(id, name),
	vars_measures %>% filter(length(traits) > 0) %>% unnest())
```

Add cultivar information so we can plot by ecotype

```{r}
traitdata <- (traitdata
	%>% rename(
		variable_name = name,
		cultivar_id = trait.cultivar_id,
		site_id = trait.site_id,
		mean = trait.mean)
	%>% mutate(date = as.Date(trait.date))
	%>% left_join(cultivars, by = "cultivar_id"))
```

```{r}
(ggplot(
	traitdata %>% filter(variable_name == "canopy_height"),
	aes(date, mean, group = site_id * cultivar_id))
	+ geom_line()
	+ facet_wrap(~ecotype)
	+ xlab("Date")
	+ ylab("Canopy height, cm"))

(ggplot(
	(traitdata
		%>% filter(variable_name == "NDVI")
		%>% mutate(emphasize = (ecotype != "RIL"))), # to reduce overplotting
	aes(date, mean, color = ecotype, group = site_id*cultivar_id, alpha = emphasize))
	+ geom_line()
	+ scale_alpha_discrete(guide = FALSE)
	+ theme(legend.position = c(0.1, 0.9)))
```
