---
title: "Plot level data from the field scanner in Maricopa, AZ"
author: "David LeBauer, Chris Black"
date: "`r Sys.Date()`"
output: md_document
---
```{r traits-03-mac-traits-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(traits)
library(rgeos)
library(sp) # for implicitly called rbind.SpatialPolygons method
library(leaflet)


options(betydb_key = readLines('~/.betykey', warn = FALSE),
        betydb_url = "https://terraref.ncsa.illinois.edu/bety/",
        betydb_api_version = 'beta')
```


## Plots


query sites for season 2

```{r traits-03-query-mac-sites}
sites <- betydb_query(
	table = "sites",
	city = "Maricopa", sitename = "~Season 2 range", limit = "none")
```

plot polygons on a map

```{r traits-03-map-mac-polygons}
site_bounds <- (sites
	%>% rowwise()
	%>% do(boundaries = readWKT(text = .$geometry, id = .$id)))

site_bounds <- do.call('rbind', site_bounds$boundaries)
names(site_bounds) <- sites$sitename

rgb20170509 <- tiff::readTIFF("/data/terraref/sites/ua-mac/Level_1/fullfield/2017-05-09/stereoTop_fullfield_10pct.tif")
leaflet() %>% 
  addTiles() %>% 
  addPolygons(data=site_bounds, popup = sites$sitename)
```

## Cultivars


```{r traits-03-mac-cultivars}
cultivars <- betydb_query(
  table = "cultivars", limit = "none") %>% 
  rename(cultivar_id = id)

cultivars_traits <- cultivars %>% 
  group_by(cultivar_id) %>% 
  do(trait_id = betydb_record(id = .$cultivar_id, table = "cultivars"
  )$traits$trait.id)	%>% 
  unnest()

sites_traits <- sites %>% 
  group_by(id) %>% 
  do(trait_id = betydb_record(id = .$id, table = "sites"
  )$traits$trait.id) %>% 
  unnest()

sites <- sites  %>% 
  left_join(sites_traits, by = "id") %>% 
  left_join(cultivars_traits, by = "trait_id") %>% 
  left_join(cultivars, by = "cultivar_id") %>% 
  unique()

```


## Time series of canopy cover, height, NDVI

First look up variables by name. Let's look for measurements related to canopy size:

```{r traits-03-height-cover-ndvi}
variables <- betydb_query(
		table = "variables", name = "~^(NDVI|canopy_height|canopy_cover|)$")

variables %>% 
  select(id, name, units, n_records = `number of associated traits`)
```

Exercise: Why are there two variables named canopy_height, and what database fields should you examine to decide which one you want?

Now retrieve all available measurements for each variable.

```{r traits-03-get-variables}

vars_measures <- (variables
	%>% group_by(id, name)
	%>% do(traits = betydb_record( # Get full trait list by variable ID
		id = .$id, table = "variables")$traits))

# Only needed if some variables may contain zero traits
# If none are empty, can just do `vars_measures %>% unnest()`
traitdata <- left_join( # ensures we keep a blank row for any variables with no trait info
	vars_measures %>% select(id, name),
	vars_measures %>% filter(length(traits) > 0) %>% unnest())
```

Add cultivar information so we can plot by ecotype

```{r traits-03-cultivar-info}
traitdata <- (traitdata
	%>% rename(
		variable_name = name,
		cultivar_id = trait.cultivar_id,
		site_id = trait.site_id,
		mean = trait.mean)
	%>% mutate(date = as.Date(trait.date))
	%>% left_join(cultivars, by = "cultivar_id"))
```

```{r traits-03-plots}
(ggplot(
	traitdata %>% filter(variable_name == "canopy_height"),
	aes(date, mean, group = site_id * cultivar_id))
	+ geom_line()
	+ facet_wrap(~ecotype)
	+ xlab("Date")
	+ ylab("Canopy height, cm"))

(ggplot(
	(traitdata
		%>% filter(variable_name == "NDVI")
		%>% mutate(emphasize = (ecotype != "RIL"))), # to reduce overplotting
	aes(date, mean, color = ecotype, group = site_id*cultivar_id, alpha = emphasize))
	+ geom_line()
	+ scale_alpha_discrete(guide = FALSE)
	+ theme(legend.position = c(0.1, 0.9)))
```
