# Danforth Indoor Phenotype Analysis

```{r 02-setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(ggthemes)
theme_set(theme_bw())

if(packageVersion("traits") == '0.2.0'){
  devtools::install_github('ropensci/traits')
}

library(traits)
```


## Connect to the TERRA REF Trait database

Unlike the first two tutorials, now we will be querying real data from the public TERRA REF database. So we will use a new URL, https://terraref.ncsa.illinois.edu/bety/, and we will need to use our own private key.

```{r terraref-connect-options}

options(betydb_key = readLines('.betykey', warn = FALSE),
        betydb_url = "https://terraref.ncsa.illinois.edu/bety/",
        betydb_api_version = 'v1')
```
### Query data from the Danforth Phenotyping Facility

First we will use the generic search to query the output from the Lemnatec indoor phenotyping system at the Danforth Center in St. Louis, MO.

```{r query-danforth, message = FALSE}
danforth_sorghum <- traits::betydb_query(
#  sitename = 'Danforth Plant Science Center Bellweather Phenotyping Facility',
  trait = 'sv_area',
  limit = 'none')

```

To get the equivalent query via the web interface, you can construct the following URL.   

```{r api-query-in-browser}
search_url <- paste0(options()$betydb_url,
                     '/api/beta/search?',
                     'scientificname=Sorghum+bicolor',
                     '&sitename=Danforth+Plant+Science+Center+Bellweather+Phenotyping+Facility',
                     '&key=', options()$betydb_key,'&limit=none')
print(gsub(options()$betydb_key, 'secretkey', search_url))
```

Once you learn how to write a query using the url API, you can use it to open in your browser to see (but you may need to grant permission)


```{r open-api-url, eval=FALSE}
browseURL(search_url)
```


Note that this returns a lot of columns. Lets take a look at what they mean

```{r all-columns}

colnames(danforth_sorghum)

```

Which columns do we want?



```{r danforth-sorghum-cleanup}
danforth_sorghum <- danforth_sorghum %>% 
  dplyr::select(date, scientificname, cultivar, entity, trait, mean, units) %>% 
  mutate(label = paste0(trait, ifelse(units == '', '', paste0(' (', units, ')'))), date = lubridate::ymd(date), genotype = cultivar)

head(danforth_sorghum)

```

### Summary of Data from the Danforth Phenotyping Facility

```{r phenotype_counts, fig.width=8, fig.height=4}


ggplot(data = danforth_sorghum) + 
  geom_histogram(aes(x = mean), bins = 100) + 
  facet_wrap(~label, scales = 'free') +
  ggthemes::theme_tufte()


```

### Growth rate over time

```{r danforth-phenotypes, fig.width=8, fig.height=4, message = FALSE}

ggplot(data = danforth_sorghum, aes(x = date, y = mean, color = cultivar)) +
 # geom_line(aes(group = entity), size = 0.1) +
  geom_smooth(aes(group = cultivar)) +
  geom_point(alpha = 0.4, size = 0.1) +
  facet_wrap(~label, scales = 'free_y') +
  labs(color = 'Cultivar') + ggthemes::theme_few() 

```

### Your turn
1. Compute phenotypes for each cultivar
2. An 'entity' is a replicate. 
   * How many entities are there? 
   * How many entities per cultivar? 
   * Did they all make it through the entire growing season?
