# Plot level data from the field scanner in Maricopa, AZ

```{r traits-05-mac-traits-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(traits)
library(rgeos)
library(sp) # for implicitly called rbind.SpatialPolygons method
library(leaflet)


options(betydb_key = readLines('.betykey', warn = FALSE),
        betydb_url = "https://terraref.ncsa.illinois.edu/bety/",
        betydb_api_version = 'v1')
```


## Plots


First, query the plots for Season 2. The simple way to use this is based on the fact that the plot names at Maricopa contain the season. 

```{r traits-05-query-mac-sites, echo = TRUE, message = FALSE}
sites <- betydb_query(
	table = "sites",
	city = "Maricopa", sitename = "~Season 2 range", limit = "none")
```

A more robust (but complicated way) would be to query the experiments and experiments_sites tables. But we will leave that as an exercise for the ambitious user.


### Plot Season 2 plots

```{r traits-05-map-mac-polygons, echo = TRUE}

site_bounds <- sites %>%
  rowwise() %>%
  do(boundaries = readWKT(text = .$geometry, id = .$id))


site_bounds <- do.call('rbind', site_bounds$boundaries)
#names(site_bounds) <- sites$sitename


#rgb20170522 <- raster::raster("/data/terraref/sites/ua-mac/Level_1/fullfield/2017-05-22/fullfield_L1_ua-mac_2017-05-22_rgb_thumb.tif", resolution = c(9.32174859971092e-10, 7.97612621550367e-10))

#image(rgb20170522)


leaflet() %>% 
  addTiles() %>% 
#  addRasterImage(rgb20170522) %>% 
  addPolygons(data=site_bounds, popup = sites$sitename)
```

```{r}

## Cultivars

```

```{r traits-05-mac-cultivars}
#cultivars <- betydb_query(
#  table = "cultivars", limit = "none") %>% 
#  rename(cultivar_id = id)


# getting HTTP 504 - Gateway timeout
# comment this chunk out
#traits <- traits <- betydb_search(
#  Season = "~Season 4", 
#  include_unchecked = 'true',
#  limit = "none") %>% 
#  rename(trait_id = id)
```

```{r}
## Time series of canopy cover, height, NDVI

#First look up variables by name. Let's look for measurements related to canopy size:

```

```

```{r traits-05-height-cover-ndvi}
#variables <- betydb_query(
#		table = "variables", name = "~^(NDVI|canopy_height|canopy_cover|)$") # a tilde ~ can be used to partially match a string #the tilde is used in this query to get variable names that contain either 'NDVI', 'canopy_height', or 'canopy_cover'

#variables %>% 
#  select(id, name, units, n_records = `number of associated traits`)
```

```{r}

#Exercise: Why are there two variables named canopy_height, and what database fields should you examine to decide which one you want?

```

```{r traits-05-get-variables-comment}

#Now retrieve all available measurements for each variable.

```

```{r traits-05-get-variables} 
#getting HTTP 504 -- Gateway Timeout when running this chunk
#comment out this chunk

#vars_measures <- (variables
#	%>% group_by(id, name)
#	%>% do(traits = betydb_record( # Get full trait list by variable ID
#		id = .$id, table = "variables")$traits))

# Only needed if some variables may contain zero traits
# If none are empty, can just do `vars_measures %>% unnest()`
#traitdata <- left_join( # ensures we keep a blank row for any variables with no trait info
#	vars_measures %>% select(id, name),
#	vars_measures %>% filter(length(traits) > 0) %>% unnest())
```

```{r traits-05-cultivar-info-comment}

#Add cultivar information so we can plot by ecotype

```

```{r traits-05-cultivar-info}
#must comment out since chunk refers to traitdata object from above (that could not be created)
#traitdata <- (traitdata
#	%>% rename(
#		variable_name = name,
#		cultivar_id = trait.cultivar_id,
#		site_id = trait.site_id,
#		mean = trait.mean)
#	%>% mutate(date = as.Date(trait.date))
#	%>% left_join(cultivars, by = "cultivar_id"))
```

```{r traits-05-plots}
#(ggplot(
#	traitdata %>% filter(variable_name == "canopy_height"),
#	aes(date, mean, group = site_id * cultivar_id))
#	+ geom_line()
#	+ facet_wrap(~ecotype)
#	+ xlab("Date")
#	+ ylab("Canopy height, cm"))

#(ggplot(
#	(traitdata
#		%>% filter(variable_name == "NDVI")
#		%>% mutate(emphasize = (ecotype != "RIL"))), # to reduce overplotting
#	aes(date, mean, color = ecotype, group = site_id*cultivar_id, alpha = emphasize))
#	+ geom_line()
#	+ scale_alpha_discrete(guide = FALSE)
#	+ theme(legend.position = c(0.1, 0.9)))
```
