---
title: "Accessing TERRA REF data using the brapi R package"
author: "Reinhard Simon"
author: "David LeBauer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

# Objective

Here demonstrate the use of the brapi R package to query data from the TERRA REF traits database (terraref.org/bety) using the Breeder's API (BrAPI).

What is BrAPI?

Data are public and no login credentials are needed.

```{r, message=TRUE, warning=TRUE}

library(brapi)
terraref <- ba_db()$terraref
print(terraref)


# show verbose server feedback
ba_show_info(TRUE)

```

## Listing available calls

The BrAPI specification does not require all endpoints to be implemented, and TERRA REF provides a subset of endpoints focused on genotypes, experimental metadata, and phenotypes. The `ba_calls()` function lists the functionality supported by the server.

```{r}
z <- ba_calls(terraref)
```

## Function Arguments

* `con`: Always the first argument, provides database connection information as a list. To query terraref, we will use `con = terraref`, which we returned above from `ba_db()$terraref` (try `print(ba_db())` to see some of the other crop databases that you can query).
* `rclass`: the last argument is always the class of object returned. The default type is a 'tibble', althought you can also request `data.frame`, `json`, or `list`.
* Other parameters are of class 'character'. Exceptions are: the con parameter is always a list; the parameters 'page' and 'pageSize' if applicable are integers. For details see individual functions.

## Getting phenotypic data

The brapi models trial data in a three layer hierarchy: a) breeding program which has b) trials that c) may consist of one or more studies at one or more locations. A study at one location is also often referred to as a fieldbook.

### Which breeding programs are there?

```{r}
ba_crops(terraref)
```


### Which studies are there?

```{r}
ba_studies_search(sp_base, programDbId = "140")
```

### Get a study (or fieldbook)

```{r, message=FALSE, warning=FALSE}
# Currently not working!!!
#dt = ba_studies_table(sp_base, 
#                      studyDbId = "151")
```

```{r, echo=FALSE}
#library(DT)
#datatable(
#  dt,
#  options=list(pageLength = 5, scrollX = TRUE)
#  )
```











