# A Simulated Phenotype Dataset

```{r include = FALSE}
library(traits)
knitr::opts_chunk$set(echo = FALSE, cache =FALSE)
library(ggplot2)
library(ggthemes)
library(GGally)
theme_set(theme_bw())
library(dplyr)
library(httr)
library(labeling)
library(highr)
library(lubridate)
library(tidyr)
.libPaths('~/R/library')
```


To explore the potential of phenotyping data as well as to prepare for large datasets, we have simulated the type of data that might be observed by daily scans by drone or robot. These data are freely accessible, and thus useful for teaching and exploration that does not require access.

## Methods: 

### The design of a simulated dataset


We simulated 500 genotypes across 12 sites and five years using a mechanistic model. 
This model was developed to simulate the physiology and biophysics of Miscanthus, which is another bioenergy grass with C4 photosynthesis, [BioCro, Miguez et al, 2009](github.com/ebimodeling/biocro). We have further added some noise to the output. [The code used to run the model and add noise is on GitHub](https://github.com/ebimodeling/biocro_regional/edit/master/vignettes/regional_pecan_workflow.Rmd).
The model computes carbon uptake at hourly time steps.
The 'genotypes' are based on five-hundred quasi-random parameterizations of a biophysical crop model.

_Note that these data sets contain numerical artifacts and scientific reinterpretations for illustrative purpose._

All of these simulated datasets are released with an unrestrive [copyright](https://creativecommons.org/publicdomain/zero/1.0/).  This means you can copy, modify, and share the data.  Please keep in mind that the data sets are not production quality - they have been developed solely to inspire and solicit feedback.


### Design of Simulation Experiment

500 Sorghum lines grown at each of three sites, four blocks per site, along a N-S transect in Illinois over five years (2021-2025). 

#### Time Span (2021-2025) 

While the climate data are derived from historic data, the years have been changed to reinforce the fact that these are not real data. The years have been chosen to select climate extremes. Two years were dry, two were wet, and one was average.

| year | drought index |
|-----|-----|
| 2021 | wet |
| 2022 | dry |
| 2023 | normal |
| 2024 | wet |
| 2025 | dry |


#### Sites

These are approximate locations used to query the meteorological and soil data used in the simulations. 
 
| site name | latitude | longitude | 
|-------|------|------|
| north | 42.0 | -88.5 |
| central | 40.0 | -88.5 |
| south | 37.0 | -88.5 |

Each site has four replicate fields: A, B, C, D. This simulated dataset assumes each field within a site has similar, but different meteorology (e.g., as if they were all in the same county). 


#### Genotypes

Two-hundred and twenty-seven lines were grown at each site. Each line is identified by a unique integer in the range [9915:10141]

#### Phenotypes

The phenotypes associated with each genotype is in the file `phenotypes.csv`. 

These 'phenotypes' are used as input parameters to the simulation model. We often refer to these as 'traits' (as opposed to biomass or growth rates, which are states and proceses). In this example, we assume that 'phenotypes' are time-invariant.


| variable_id|name                           |standard_name                                                 |units                          |Description                                                                                                                                                                                                      |
|-----------:|:------------------------------|:-------------------------------------------------------------|:------------------------------|:----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
|          |genotype                 |                            |               |genetically and phenotypically distinct line |         
|          |Vmax                 |                            | umol m-2 s-1              | maximum carboxylation of Rubisco according to the Collatz model  |      
|          38|cuticular_cond                 |conductance_of_fully_closed_stomata                           |umol H2O m-2 s-1               |leaf conductance when stomata fully closed                                                                                                                                                                       |
|          15|SLA                            |specific_leaf_area                                            |m2 kg-1    |Specific Leaf Area|
|          39|quantum_efficiency             |mole_ratio_of_carbon_dioxide_to_irradiance_in_leaf  |fraction   |see Farqhuar model |
|          18|LAI                            |leaf_area_index          |m2 leaf m-2 ground |Leaf Area Index |
|          31|c2n_leaf |mass_ratio_of_carbon_to_nitrogen_in_leaf   |ratio  |C:N ratio in leaves|
|         493|growth_respiration_coefficient |respiration_coefficient_for_growth  |mol CO2 / mol net assimilation |amount of CO2 released due to growth per unit net photosynthesis  |
|           7|leaf_respiration_rate_m2       |respiration_rate_per_unit_area_in_leaf                        |umol CO2 m-2 s-1               |Not really ""dark respiration"" Often this is respiration that occurs in the light. Date and time fields ""should"" identify pre-dawn (nightime/dark) leaf resp vs the Rd that comes from a A-Ci or A-PPFD curve |
|           4|Vcmax                          |rubisco_carboxylation_rate_in_leaf_assuming_saturated_rubp    |umol CO2 m-2 s-1               |maximum rubisco carboxylation capacity  |
|         404|stomatal_slope.BB              |stomatal_slope_parameter_assuming_ball_berry_model            |ratio      |slope parameter for Ball-Berry Model of stomatal conductance |
|           5|Jmax   |electron_transport_flux_in_thylakoid_assuming_saturated_light |umol photons m-2 s-1   |maximum rate of electron transport  |
|         492|extinction_coefficient_diffuse |extinction_coefficient_for_diffuse_light_in_canopy            |                               |canopy extinction coefficient for diffuse light                                                                                                                                                                  |

#### Simulated Sensor Data

This dataset includes what a sensor might observe, daily for five years during the growing season.

| variable_id|name                           |standard_name       |units                          |Description                                                                                                                                                                                                      |
|-----------:|:------------------------------|:---------------|:------------------------------|:----------------|
|            | sitename                     |                                                              |                        |  Name of site |
|            | plotid                               |                                                              |                        | experimental replicate plot |
|            |year                             |                                                              |                        | |
|            | date                           | | YYYY-MM-DD         | |
|            |  Stem                             |  stem_biomass_content  |                        | Mg / ha |
|            |  Leaf                             |  leaf_biomass_content  |                        | Mg / ha |
|            |  Root                             |  root_biomass_content  |                        | Mg / ha |
|            |  Rhizome                             |  rhizome_biomass_content  |                        | Mg / ha |
|          18|LAI                            |leaf_area_index   |ratio             |  Leaf Area Index is the ratio of leaf area to ground area |
|          |NDVI                            |normalized_difference_vegetation_index                                               |ratio             | commonly used vegetation index |
|            | Height                               |   canopy_height | m                        | |




### Accessing the TERRA Simulated Data Database


```{r set-options}

options(betydb_key = readLines('.betykey_public', warn = FALSE),
        betydb_url = "https://terraref.ncsa.illinois.edu/bety-test/",
        betydb_api_version = 'v1')


```


```{r}

sorghum_sla <- betydb_query(table = 'search',
                            trait = "SLA",
                            genus = 'Sorghum',
                            limit = 500)

```


Now we have queried all `r nrow(sorghum_sla)` of the Specific Leaf Area records, we can see how this trait varies within the simulated genotypes. 


```{r sla-fig, fig.height=2, fig.width=6}
ggplot(sorghum_sla) + 
  geom_histogram(aes(mean), binwidth = 1)
```

## Your turn: query the list of available traits from the variables table

```{r query-traits, message = FALSE}



trait_list <- c("Vcmax", "c2n_leaf", "cuticular_cond", "SLA", "quantum_efficiency", "leaf_respiration_rate_m2", "stomatal_slope.BB", "Jmax", "chi_leaf", "extinction_coefficient_diffuse")

variables <- betydb_query(table = 'variables',
                          limit = 'none')


knitr::kable(variables %>% 
               filter(name %in% trait_list) %>% 
               select(name, description, units))
        
```

### Select and evaluate traits

These traits are not time series, each of the ~500 genotypes is associated with a single value for each trait. This is different from the time series of LAI that we saw in the previous exercise or the biomass data that we will look at below. 

```{r traits-sel, message = FALSE}

traits_list <- list()
for(trait in trait_list){
  traits_list[[trait]] <- betydb_query(table = 'search',
                                  trait = trait,
                                  limit = 'none')
}

traits <- bind_rows(traits_list)
  
```



Lets summarize the trait data as histograms:

```{r phenotype-histograms, fig.width = 6, fig.height = 6}
ggplot(data = traits) +
  geom_histogram(aes(mean), bins = 22) +
  facet_wrap(~trait, scales = 'free', ncol = 2) + 
  theme_minimal()

```


### Plant Growth time series

Lets now select a single cultivar and view its growth over time across many sites. There are time series of NDVI, canopy height, leaf area index, leaf biomass, root biomass, and stem biomass. 

```{r}
timeseries_traits <- c("canopy_height", "leaf_biomass", "root_biomass", "stem_biomass", "LAI", "NDVI")

knitr::kable(variables %>% 
               filter(name %in% timeseries_traits) %>% 
               select(name, description, units))
```

```{r all_sorghum, cache = TRUE, results = 'hide'}


site_id <- betydb_query(table = 'sites', sitename = "Central IL Plot D")$id

for(t in c('canopy_height', 'stem_biomass', 'LAI', 'NDVI')){
  
  variable_id <- betydb_query(table = 'variables', name = t)$id

  system.time(
    sorghum_ts <-  betydb_query(table = 'traits',
                                limit = 'none',
                                variable_id = variable_id,
                                date_year ="1022",
                                site_id = site_id)
  )
  
#  save(sorghum_ts, file = paste0('data/sorghum_',t,'.RData'))
}

```


This is how you can query a time series of sorghum height data for the Northern IL site. 

```{r query-sorghum-height, echo = TRUE}
#sorghum_height <-  betydb_query(table = 'search',
#                                trait = 'canopy_height',
#                                year = 1022,
#                                site = "~Northern IL",
#                                limit = 'none')

#save(sorghum_height, file = 'traits/sorghum_height.RData')
```

However, with almost 200k rows it currently takes 40 minutes to query (this is a limitation of the API). For the purposes of this tutorial, we will use a cached copy of the dataset.


```{r 10-sim-sorg-plot, message = FALSE}
load('traits/sorghum_height.RData')

s <- sorghum_height %>% 
  mutate(day = lubridate::yday(raw_date), 
         year = lubridate::year(raw_date),
         genotype = as.factor(cultivar),
         height = mean)


height_max <- s %>% group_by(genotype) %>% summarise(height_rank = max(height))

ggplot(data = s, aes(x = day, y = height), size = 0.25, se = FALSE) + 
  geom_smooth(aes(group = genotype), alpha = 0.1)

        
```



### Trait Correlations

Now lets look at a 'pairs' plot to see if there is any covariance among the traits:


First, lets rearrange the data from 'long' to 'wide' format. We will also take this chance to rename the 'cultivar' field to 'genotype'.

```{r 10_traits_wide, echo = TRUE}

traits_wide <- traits %>%  
  select(genotype = cultivar, trait, mean) %>%
  tidyr::spread(key = trait, value = mean)

```

Now, lets create a variable called `max_height` 

```{r max_height, echo = TRUE}
# create the variable max height
max_height <- s %>% 
  group_by(genotype) %>% 
  summarise(max_height = max(height))

```

Now, join the traits data frame with the new max_height data frame trait data we will merge the two data frames on the `genotype` field.

```{r join_traits_height, echo = TRUE, warning = FALSE}

traits_height <- traits_wide %>% left_join(max_height, by = 'genotype')

```
Which traits are related to height? We can discover this in a few way, for example, a pairs plot that shows correltations:

```{r trait_pairs, fig.height = 8, fig.width = 8, warning = FALSE}
ggpairs(traits_height %>% select(-genotype),  
        lower = list(continuous = 'density'),
        upper = list(continuous = 'cor'),
        axisLabels = 'show') +  
  theme(text = element_text(size=6))

```

What we can see in this _simulated_ data set is that there is no covariance among traits. However, many of the traits are correlated to the maximum height. Is this realistic? What parameters do you think should be correlated? We will re-visit this with real data in a later tutorial. 

We can also use linear regression to evaluate the relationship between traits and height:

```{r trait-anova}
height_model <- lm(max_height ~ #. -genotype,
                     c2n_leaf + chi_leaf + cuticular_cond + extinction_coefficient_diffuse +
                     Jmax + leaf_respiration_rate_m2 + quantum_efficiency + SLA, + Vcmax, 
                   data = traits_height)
summary(height_model)


```
