---
title: "First Walkthrough Notes"
author: "Kristina Riemer"
output: github_document
urlcolor: blue
---

## Video 1: Objectives + TERRA REF Overview + Getting Started

have time to install everything/open up, put up instructions on screen from email

#### Objectives
These walkthroughs are a hands-on introduction to how to use data produced by TERRA REF project. 

The intended audience is community members and anyone interested in these data. 

These live walkthroughs will be turned into video tutorials that can be viewed at any time. They will be put on YouTube, and I'll provide links once they're ready. 

#### Project overview
TERRA REF is a project funded by ARPA-E to gather phenomic and genomic data on plant crops. An extensive series of sensors are used to collect trait and image data on outdoor experimental plots. Weather stations also collect weather data. And crop genomes are being sequenced, with these genomic data being provided. 

These tutorials are to make these data accessible and useable. I will provide an overview of how to use computational tools to get these data, and combine, plot, and analyze them. 

We have written online tutorials [here](https://terraref.github.io/tutorials/). These walkthroughs and videos complement those materials, which are growing and improving continually. 

#### VICE setup and use
Launching TERRA REF app on VICE: 

1. Google "Cyverse" and click link to website
2. Select Products tab
3. Under "Discovery Environment", select Launch button
4. Hit "Log in with your CyVerse ID" button
5. Log in with CyVerse credentials (username and password)
6. Select "Apps", type in "terra", select "TERRA REF Rstudio 3.6.0", and "Use App"
7. Click "Launch Analysis", hit Analysis button, and select arrow button in Analyses window
8. Wait patiently
9. Sign into RStudio using username: rstudio and password: rstudio1

Reopen running TERRA REF instance: 

1. Google "Cyverse" and click link to website
2. Select Products tab
3. Under "Discovery Environment", select Launch button
4. Hit "Log in with your CyVerse ID" button
5. Log in with CyVerse credentials (username and password)
6. Open Analyses window and hit arrow button to open up RStudio

Using R language within RStudio. Just like you'd use on your own computer when you open up RStudio except because this is within VICE, all the packages we need are already installed. 

[if you're not ready, join at weather part]

## Video 2: Downloading Trait Data

Use `traits` R package to download TERRA REF trait data. Package developed by rOpenSci more generally for downloading trait data. Implemented for several other trait databases in addition to TERRA REF. Package is on CRAN. 

```{r}
library(traits)
```

First set some global options that will apply across all instances of using this package's function. Limits redundant typing. 

Specify: 

- Where from internet to download from (NCSA is where TERRA REF data is)
- Want to use most recent version of API for data
- Using public API key to access data, will show later how to access your own API key to get to more data

```{r}
options(betydb_url = "https://terraref.ncsa.illinois.edu/bety/", 
        betydb_api_version = 'beta', 
        betydb_key = '9999999999999999999999999999999999999999')
```

Function from `traits` is `betydb_query`. Getting first 1000 rows of data from fourth season. Can get all by not setting `limit` argument, but it takes a while. 

```{r}
season_4 <- betydb_query(sitename = "~Season 4", 
                         limit = 1000)
```

Each row is trait observation. Lots of info, including location (`lat`, `lon`), species, time, type of `trait`, and value (`mean`). 

## Video 3: Available Traits

Look at which traits are available. Using popular data cleaning R package `dplyr`. 

Take dataframe with trait observations, return all the unique values from trait column with `distinct` and print them all out. 

```{r}
library(dplyr)
season_4 %>% 
  distinct(trait) %>% 
  print(n = Inf)
```

Focus on one trait, canopy_height, from more recent season. 

```{r}
canopy_height <- betydb_query(trait     = "canopy_height", 
                              sitename  = "~Season 6",
                              limit     =  250)
```

## Video 4: Plot Trait Data

Want to plot canopy height across time using `raw_date`. Column is in wrong format, so first use a function from `lubridate` package to transform. 

Add this as new column using `mutate` from `dplyr`. 

```{r}
library(lubridate)
canopy_height <- canopy_height %>% 
  mutate(formatted_date = ymd_hms(raw_date))
```

Use common R package `ggplot` to plot these data. Specify dataframe, then what axes should be with `aes` argument. Use scatter plot function. 

```{r}
library(ggplot2)
ggplot(data = canopy_height, aes(x = formatted_date, y = mean)) +
  geom_point()
```

Add better axis labels. Can find canopy_height units from `units`. 

```{r}
ggplot(data = canopy_height, aes(x = formatted_date, y = mean)) +
  geom_point() +
  labs(x = "Date", y = "Plant height (cm)")
```

Can do same plot across all the traits in the original dataframe from fourth season. Do same column time transformation. 

```{r}
season_4 <- season_4 %>%
  mutate(formatted_date = ymd_hms(raw_date))
```

Create plot same as for canopy_height, but add in `facet_wrap` to produce one per unique trait type. 

Some data only take at one or a few times of year. 

```{r}
ggplot(data = season_4, aes(x = formatted_date, y = mean)) +
  geom_point() +
  facet_wrap(~trait, scales = "free_y")
```

## Video 5: Access API Key

Each beta user will have their own unique key. Do not share these with anyone. 

Find the key: 

1. Log into betydb.org
2. Go to data/users
3. See your account there with API key listed

## Video 6: Downloading Weather Data

There's no R package for downloading TERRA REf weather data. 

Data is on a platform called Clowder. It's in format called JSON. Use an R package that works with JSON data. `fromJSON` is function from that package that pulls data down and turns into R's data frame format. 

```{r}
library(jsonlite)
```

First specify URL where data comes from. Parts are from NCSA Clowder, using API and geostreams, then weather station with stream_id and since and until for date range. 

Then read into `fromJSON`. This gets all available weather data for 2017. 

```{r}
weather_url <- "https://terraref.ncsa.illinois.edu/clowder/api/geostreams/datapoints?stream_id=46431&since=2017-01-02&until=2017-01-31"
weather <- fromJSON(weather_url, flatten = FALSE)
```

Subset of data is called properties, which contains what we want. Pull that out. Rows are weather observations, and columns are different weather variables. 

```{r}
weather_obs <- weather$properties
```

Want to get date, which is in original weather data frame. Take that column, change into correct format, and add to observations data frame. 

```{r}
weather_obs <- weather$properties %>% 
  mutate(formatted_date = ymd_hms(weather$end_time))
```

## Video 7: Plot Single Weather Variable

Want to look at air temperature across. Plot these data like we did before for trait data. Turns out these data are only for month of January in 2017. 

```{r}
ggplot(data = weather_obs, aes(x = formatted_date, y = air_temperature)) +
  geom_point() +
  labs(x = "Date", y = "Temperature (K)")
```

## Video 8: Plot All Weather Variables

Want to plot all weather variables across time to see when they are available and what their values are like. 

Need to rearrange the dataframe to do this in ggplot. Data are currently in wide format, will change into long format. That is, we want to have one column with the name of the weather variable and another column with the value for that variable. 

First remove a couple of unneeded columns. Then turn variable header names into a new column `variable` and their values in a new `value` column. Last, the date column is included. 

```{r}
library(tidyr)
weather_obs_long <- weather_obs %>% 
  select(-source, -source_file) %>% 
  gather(variable, value, -formatted_date)
```

All observations are plotted across time then. `facet_wrap` is used to create one plot per weather variable. 

```{r}
ggplot(data = weather_obs_long, aes(x = formatted_date, y = value)) +
  geom_point() +
  facet_wrap(~variable, scales = "free_y") +
  labs(x = "Date", y = "Weather variable")
```

In addition to temperature, there's wind speed, rainfall (not much in Arizona), humidity, and radiation. 
