---
title: "First Walkthrough"
author: "Kristina Riemer"
output: github_document
urlcolor: blue
---

### Video 1: Objectives + TERRA REF Overview

TBD

### Video 2: Downloading Trait Data

Use `traits` R package to download TERRA REF trait data. Package developed by rOpenSci more generally for downloading trait data. Implemented for several other trait databases in addition to TERRA REF. Package is on CRAN. 

```{r}
library(traits)
```

First set some global options that will apply across all instances of using this package's function. Limits redundant typing. 

Specify: 

- Where from internet to download from (NCSA is where TERRA REF data is)
- Want to use most recent version of API for data
- Using public API key to access data, will show later how to access your own API key to get to more data

```{r}
options(betydb_url = "https://terraref.ncsa.illinois.edu/bety/", 
        betydb_api_version = 'beta', 
        betydb_key = '9999999999999999999999999999999999999999')
```

Function from `traits` is `betydb_query`. Getting first 1000 rows of data from fourth season. Can get all by not setting `limit` argument, but it takes a while. 

```{r}
season_4 <- betydb_query(sitename = "~Season 4", 
                         limit = 1000)
```

Each row is trait observation. Lots of info, including location (`lat`, `lon`), species, time, type of `trait`, and value (`mean`). 

### Video 3: Available Traits

Look at which traits are available. Using popular data cleaning R package `dplyr`. 

Take dataframe with trait observations, return all the unique values from trait column with `distinct` and print them all out. 

```{r}
library(dplyr)
season_4 %>% 
  distinct(trait) %>% 
  print(n = Inf)
```

Focus on one trait, canopy_height, from more recent season. 

```{r}
canopy_height <- betydb_query(trait     = "canopy_height", 
                              sitename  = "~Season 6",
                              limit     =  250)
```

### Video 4: Plot Trait Data

Want to plot canopy height across time using `raw_date`. Column is in wrong format, so first use a function from `lubridate` package to transform. 

Add this as new column using `mutate` from `dplyr`. 

```{r}
library(lubridate)
canopy_height <- canopy_height %>% 
  mutate(formatted_date = ymd_hms(raw_date))
```

Use common R package `ggplot` to plot these data. Specify dataframe, then what axes should be with `aes` argument. Use scatter plot function. 

```{r}
library(ggplot2)
ggplot(data = canopy_height, aes(x = formatted_date, y = mean)) +
  geom_point()
```

Add better axis labels. Can find canopy_height units from `units`. 

```{r}
ggplot(data = canopy_height, aes(x = formatted_date, y = mean)) +
  geom_point() +
  labs(x = "Date", y = "Plant height (cm)")
```

Can do same plot across all the traits in the original dataframe from fourth season. Do same column time transformation. 

```{r}
season_4 <- season_4 %>%
  mutate(formatted_date = ymd_hms(raw_date))
```

Create plot same as for canopy_height, but add in `facet_wrap` to produce one per unique trait type. 

Some data only take at one or a few times of year. 

```{r}
ggplot(data = season_4, aes(x = formatted_date, y = mean)) +
  geom_point() +
  facet_wrap(~trait, scales = "free_y")
```

### Access API Key

Each beta user will have their own unique key. Do not share these with anyone. 

Find the key: 

1. Log into betydb.org
2. Go to data/users
3. See your account there with API key listed
